---
  output: github_document
---
  
```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%",
  dpi=300
)
```

```{r, echo=FALSE, results='asis'}
# Define your branch variable
current_branch <- system("git rev-parse --abbrev-ref HEAD", intern = TRUE)

cat(paste0("# ProCESS <a href=\"caravagnalab.github.io/ProCESS/tree/",
           current_branch,
           "\"><img src=\"man/figures/logo.png\" align=\"right\" ",
           "height=\"120\" alt=\"\" /></a>\n"))
```

<!-- badges: start -->
<!--  
  [![R-CMD-check](https://github.com/caravagnalab/ProCESS/workflows/R-CMD-check/badge.svg)](https://github.com/caravagnalab/ProCESS/actions)
  [![pkgdown](https://github.com/caravagnalab/ProCESS/actions/workflows/pkgdown.yaml/badge.svg)](https://github.com/caravagnalab/ProCESS/actions/workflows/pkgdown.yaml)
-->
<!-- badges: end -->
  
`ProCESS` stands for Programmable Cancer Evolution Spatial Simulator.
It is an R wrapper for
[CLONES](https://github.com/albertocasagrande/CLONES), a C++ tumour
evolution simulator, and provides additional plotting functions.

-----

#### Help and support

```{r, echo=FALSE, results='asis'}

site <- paste0("https://caravagnalab.github.io/ProCESS/",
               current_branch)

badge <- paste0("[![ProCESS GitHub Pages](https://img.shields.io/badge/",
           "GitHub%20Pages-",site,"-yellow.svg)]",
           "(",site,")")

cat(paste0(badge,"\n"))
```


----

### Installation

In order to install the development version of `ProCESS`, you need:

* [R and Rtools](https://cran.r-project.org)

* the R package [`pak`](https://pak.r-lib.org)

* [git](https://git-scm.com/downloads)


When the requirements have been satisfied, issue the R command:

```r
```{r, echo=FALSE, results='asis'}

default_branch <- system(paste0("git rev-parse --abbrev-ref origin/HEAD ",
                                "| cut -d'/' -f2"), intern = TRUE)

if (current_branch == default_branch) {
  location <- ""
} else {
  location <- paste0("@", current_branch)
}

cat(paste0("pak::pak(\"caravagnalab/ProCESS", location, "\")"))
```
```

-----

### A Simple Example

```{r, eval=FALSE, collapse=TRUE}
library(ProCESS)
```
```{r, echo=FALSE, message=FALSE, hide, collapse=TRUE}
devtools::load_all()
```
```{r tissue,fig.height=4, fig.width=4, fig.alt=c("Pre-sampling tissue.","Post-sampling tissue.")}
# set the seed of the random number generator for repeatability
set.seed(0)

# create a tissue simulation with two epigenetic states
sim <- TissueSimulation(epigenetic_states = c("E1", "E2"),
                        width = 300, height = 300)

# add a mutant "A" and set its species rates
sim$add_mutant("A", list(E1 = list(duplication = 2, death = 1,
                                   E2 = 0.12),
                         E2 = list(duplication = 0.8, death = 0.1,
                                   E1 = 0.12)))

# place one cell of "A" in epigenetic state "E1"
sim$place_cell("A[E1]", 150, 150)

# let the simulation evolve until the species "A[E2]" has less than 10 cells
sim$run_up_to_size("A[E2]", 10)

# add a mutant "A" and set its species rates
sim$add_mutant("B", list(E1 = list(duplication = 3, death = 1,
                                   E2 = 0.12),
                         E2 = list(duplication = 1.3, death = 1,
                                   E1 = 0.12)))

# choose a border cell in "A" and let one of its progeny mutate in "B"
sim$mutate_progeny(sim$choose_border_cell_in("A"), "B")

# let the simulation evolve until the species "A[E2]" has less than 3000 cells
sim$run_up_to_size("B[E1]", 2e4)

# plot the tissue
plot_tissue(sim)

# collect 3 samples: "Sample_A", "Sample_B", and "Sample_C"
sim$sample_cells("Sample_A", c(125, 125), c(175, 175))
sim$sample_cells("Sample_B", c(175, 175), c(225, 225))
sim$sample_cells("Sample_C", c(50, 100), c(100, 150))

# let the simulation evolve until the species "B[E1]" has less than 80k cells
sim$run_up_to_size("B[E1]", 2e4)

# plot the tissue
plot_tissue(sim)
```

```{r forest,fig.height=4, fig.width=8, fig.alt=c("Sample forest")}
# get the sample forest
sample_forest <- sim$get_sample_forest()

# plot it
plot_forest(sample_forest) %>% 
  annotate_forest(sample_forest)
```

```{r, echo=FALSE, results='asis'}

cat(paste0("\nFor more advanced usage examples, please refer to\n",
           badge))
```

-----

#### Copyright and contacts
  
- Alberto Casagrande, Computational Biology and Bioinformatics@UniUd.
- Giulio Caravagna, Cancer Data Science (CDS) Laboratory.

[![](https://img.shields.io/badge/CDS%20Lab%20Github-albertocasagrande-seagreen.svg)](https://github.com/albertocasagrande/)
[![](https://img.shields.io/badge/CDS%20Lab%20Github-caravagnalab-seagreen.svg)](https://github.com/caravagnalab/)
[![](https://img.shields.io/badge/CBB%20Lab%20webpage-https://bioinf.dimi.uniud.it/-blue.svg)](https://bioinf.dimi.uniud.it/)
[![](https://img.shields.io/badge/CDS%20Lab%20webpage-https://www.caravagnalab.org/-red.svg)](https://www.caravagnalab.org/)
