---
title: "Customizing mutation engine"
output: rmarkdown::html_vignette
description: >
  How to customize the mutation engine and its indices
vignette: >
  %\VignetteIndexEntry{Customizing mutation engine}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options:
  markdown:
    wrap: 72
bibliography: references.bib
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

> *Disclaimer:* ProCESS/CLONES internally implements the
> probability distributions using the C++11 random number
> distribution classes. The standard does not specify their
> algorithms, and the class implementations are left free
> for the compiler. Thus, the simulation output depends on
> the compiler used to compile [CLONES](https://github.com/albertocasagrande/CLONES), and because of that,
> the results reported in this article may differ from
> those obtained by the reader.

ProCESS/CLONES can simulate genomic mutations in the cells represented in
a `SampleForest` according to the specified SBS and indel mutational signatures
(see the [article about mutations](mutations.html)).
This task is achieved by the mutation engine, which uses two indices on the
reference genome to place the most probable mutations according to
signatures: the _context index_ and the _repeated sequence index_.

The function `MutationEngine()` builds the mutation engine and
these two indices.

## Customizing the mutation engine

While the most convenient way to invoke `MutationEngine()` is by using
the parameter `setup_code` (see the [article about mutations](mutations.html)),
there are cases in which a custom setup is required. In these cases,
the function `MutationEngine()` can be called by specifying the name
of the setup directory, the path or URL of the reference sequence, the
signature files, the driver mutation file, the passenger CNAs file, and
the germline data directory.

```{r}
library("ProCESS")

reference_url <- paste0("https://ftp.ensembl.org/pub/grch37/release-111/",
                        "fasta/homo_sapiens/dna/Homo_sapiens.GRCh37.",
                        "dna.chromosome.22.fa.gz")

sbs_url <- paste0("https://zenodo.org/records/15875185/files/",
                  "SBS_demo_signatures.txt")

indel_url <- paste0("https://zenodo.org/records/15875185/files/",
                    "indel_demo_signatures.txt")

drivers_url <- paste0("https://zenodo.org/records/15875185/files/",
                      "driver_mutations_hg19.csv.bz2")

passenger_cnas_url <- paste0("https://zenodo.org/records/15875185/files/",
                             "passenger_CNAs_hg19.csv.bz2")

germline_url <- paste0("https://zenodo.org/records/15875185/files/",
                       "germline_data_demo.tar.bz2")

# build a mutation engine and place all the files in the directory "Test"
m_engine <- MutationEngine(directory = "Test",
                           reference_src = reference_url,
                           SBS_signatures_src = sbs_url,
                           indel_signatures_src = indel_url,
                           drivers_src = drivers_url,
                           passenger_CNAs_src = passenger_cnas_url,
                           germline_src = germline_url)

dir.exists("Test")
```

```{r, echo=FALSE,  results='hide', message=FALSE}
unlink("Test", recursive = TRUE)
```

### The Context Index

The SBS signatures set, for every possible context
(i.e., a triplet of consecutive nucleotides), the probability for an SNV
to occur in such a context (see [@alexandrov2020repertoire]). Hence, the
function `MutationEngine()` builds
a context index of the reference sequence to place SNVs on the forest
cells. The complete context index of a sequence whose size is up to 4Gbps
takes 4 times the length of the sequence itself, both in memory and on the disk,
because 4 bytes per nucleotide are required to store the context position.
For instance, the complete context index of a human genome takes about 12 GB.
In order to avoid such a requirement in memory and disk space,
`MutationEngine()` allows sampling the reference genome contexts and
storing only some of them in the context index. This is achieved by
the optional parameter `context_sampling`, which specifies how many
occurrences of the same context must be identified before adding one of
them to the context index. The larger the number of context sampling,
the larger the context index. On the other side, the lower the number
of context sampling, the lower the number of sites in the reference
genome that can be affected by simulated mutations.
The `context_sampling` is set to 100 by default, but it can
be specified during the `MutationEngine()` call.

```{r}
m_engine <- MutationEngine(setup_code = "demo")

# get the size of the context index when `context_sampling` is 100
utils:::format.object_size(file.size("demo/context_index_100.cif"), "auto")
```

Let us rebuild the context index, sampling one context every 50.

```{r}
# building a mutation engine by using the "demo" set-up configuration
# and setting context_sampling to 50
m_engine <- MutationEngine(setup_code = "demo", context_sampling = 50)
```

The above call uses the already downloaded data in the directory `"demo"` to
build the context and produces the file `"demo/context_index_50.cif"`.

```{r}
# get the size of the context index when `context_sampling` is 50
utils:::format.object_size(file.size("demo/context_index_50.cif"), "auto")
```

### The Repeated Sequence Index

Analogously to the SBS signatures, the ID (indel) signatures set, for
some kinds of repeated sequences, the probability for an indel
to occur in them (see [@alexandrov2020repertoire]). Thus, the function
`MutationEngine()` also builds an index for the repeated sequences in the
reference.
As in the case of the SBS signatures, the ID signatures can be memory eager.
To mitigate the memory requirement, `MutationEngine()`
accepts as parameters the maximum size of the considered indel
(the parameter `max_index_size`) and the maximum number of reference positions
stored per repetition type (the parameter `max_repetition_storage`).
The larger the two parameters, the larger the repetition index.
As for building time, the larger the max_index_size, the longer it takes to 
construct the index. On the contrary, `max_repetition_storage` does not
directly affect the computation time.
By default, the two parameters are set to 50 and 500000, respectively.
However, their values can
be specified during the `MutationEngine()` call.

```{r}
# get the size of the repeated index when the default parameter values
# are used
utils:::format.object_size(file.size("demo/rs_index_50_500000.rsif"), "auto")
```

Let us rebuild the context index, sampling one context every 50.

```{r}
# building a mutation engine by using the "demo" set-up configuration
# and setting max_repetition_storage to 10,000,000
m_engine <- MutationEngine(setup_code = "demo",
                           max_repetition_storage = 10000000)
```

The above call uses the already downloaded data in the directory `"demo"` to
build the repeated sequence index and produces the file
`"demo/rs_index_50_10000000.rsif"`.

```{r}
# get the size of the context index when `context_sampling` is 50
utils:::format.object_size(file.size("demo/rs_index_50_10000000.rsif"),
                           "auto")
```
