---
title: "A four-mutants example"
output: rmarkdown::html_vignette
description: >
  Sampling a four-mutants simulation
vignette: >
  %\VignetteIndexEntry{A four-mutants example}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options:
  markdown:
    wrap: 72
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

> *Disclaimer:* ProCESS/CLONES internally implements the
> probability distributions using the C++11 random number
> distribution classes. The standard does not specify their
> algorithms, and the class implementations are left free
> for the compiler. Thus, the simulation output depends on
> the compiler used to compile [CLONES](https://github.com/albertocasagrande/CLONES), and because of that,
> the results reported in this article may differ from
> those obtained by the reader.

We build a tumour with a single mutant `A` and grow it up to
10 cells.

```{r}
library(ProCESS)

# set the seed of the random number generator for repeatability
set.seed(0)

sim <- TissueSimulation(width = 600, height = 600)

# set the delta time between two time series samples
sim$history_delta <- 10

# avoid drift
sim$death_activation_level <- 50

sim$add_mutant(name = "A", growth_rates = 0.12, death_rates = 0.05)

sim$place_cell("A", 300, 300)
sim$run_up_to_size("A", 10)
```

We add a new mutant `B` as a sub-clone of `A` and let the simulation
evolve until `B` reaches 30 cells.
```{r}
sim$add_mutant(name = "B", growth_rates = 0.145, death_rates = 0.06)
sim$mutate_progeny(sim$choose_cell_in("A"), "B")

sim$run_up_to_size("B", 30)
```

We add a further mutant `C` as a sub-clones for `A` and let the tumour
grow again up to the point at which `C` consists of 25000 cells.
```{r}
sim$add_mutant(name = "C", growth_rates = 0.15, death_rates = 0.06)
sim$mutate_progeny(sim$choose_cell_in("A"), "C")

sim$run_up_to_size("C", 25000)
```

Then, we plot the tissue and the simulation state.

```{r, fig.width=3, fig.height=3, fig.alt=c("The tissue of a tumour having three mutants: A, B, and C. Both B, and C are sub-clones of A. The three mutants have different grow rates and appears at different simulated time.")}
plot_tissue(sim)
```

```{r, fig.width=3, fig.height=3, fig.alt=c("The current ratio of cells per species over all tumour cells.")}
plot_state(sim)
```

## Tissue sampling level


We can collect two samples: "S1" and "S2".

```{r}
sim$sample_cells("S1", c(145, 230), c(215, 300))
sim$sample_cells("S2", c(350, 300), c(420, 370))
```

We can plot the tissue simulation after the sampling. We label the sampled
region to improve readability, but it is not mandatory.

```{r, fig.width=3, fig.height=3, fig.alt=c("The three-mutants tumour just after collecting two samples: S1 and S2. The former mainly contains C cells; the latter A cells.")}
plot_tissue(sim) +
  ggplot2::annotate("text", x = (145 + 215) / 2, y = (230 + 300) / 2,
                    label = "S1", parse = TRUE) +
  ggplot2::annotate("text", x = (350 + 420) / 2, y = (300 + 370) / 2,
                    label = "S2", parse = TRUE)
```

After the sampling, we can add a new mutant `D` as a sub-clone of `B`, let the
simulation continue to evolve until the sum of the cardinalities of mutants `C`
and `D` is 100000, and sample the tissue again.

```{r, fig.width=3, fig.height=3, fig.alt=c("After collecting the samples S1 and S2, a more aggressive mutant D arises from B and a further sample S3, containing both D and C cells, is collected.", "The simulation state after collecting S3.")}
sim$add_mutant(name = "D", growth_rates = 0.8, death_rates = 0.01)
sim$mutate_progeny(sim$choose_cell_in("B", c(240, 130), c(250, 150)), "D")

sim$run_until(sim$var("C") + sim$var("D") == 100000)

sim$sample_cells("S3", c(360, 80), c(430, 150))

plot_tissue(sim) +
  ggplot2::annotate("text", x = (360 + 430) / 2, y = (80 + 150) / 2,
                    label = "S3", parse = TRUE)

plot_state(sim)
```

The time-series plot represents the species cardinalities along the simulation.

```{r, fig.width=4, fig.alt=c("The time series plot represent the species cardinalities along the simulation.")}
plot_timeseries(sim)
```

The Muller plot, instead, shows the percentage of cells in each species over
the course of the simulation.

```{r, fig.width=4, fig.alt=c("The Muller plot overviews the percentage of cells in each species along the simulation evolution.")}
plot_muller(sim)
```

We can build the sample forest and plot it.

```{r, fig.width=7, fig.alt=c("The sample forest of the collected samples.")}
sample_forest <- sim$get_sample_forest()

library(dplyr)

plot_forest(sample_forest) %>%
  annotate_forest(sample_forest)
```

## Cell genetic characterisation level

We first build the mutation engine to label each cell in the sample forest.
The mutation engine pre-defined setups automatically download all the needed
files. All the pre-configured setups, but "demo" requires a COSMIC account to
download the signature files from the COSMIC site.

```{r}
# build the mutation engine according to the pre-defined setup "demo"
m_engine <- MutationEngine(setup_code="demo",
                           COSMIC_account = list(email = "foo@bar.com",
                                                 password = "#########"))

```

We must genomically characterise the tissue simulation mutants, providing
passenger mutation rates and the list of driver mutations.

```{r}
# genetically characterise the mutants
m_engine$add_mutant("A", passenger_rates = c(SNV = 2e-9, indel = 2e-9),
                    drivers = list(Mutation("22", 16085675, "GCCTCCCGA", "G"),
                                   CNA("D", "22", 22453799, 200000,
                                       allele = 1)))

m_engine$add_mutant("B", passenger_rates = c(SNV = 1e-9, indel = 2e-9,
                                             CNA = 1e-9),
                    drivers = list(SNV("22", 16050185, "C", allele = 1),
                                   CNA(type = "A", chr = "22",
                                       from = 16485130, len = 200000)))

m_engine$add_mutant("C", passenger_rates = list(SNV = 2e-8, indel = 2e-9),
                    drivers = list(SNV("22", 32786322, "G"),
                                   list("DGCR8 P26L", allele = 1)))

m_engine$add_mutant("D", passenger_rates = list(SNV = 2e-9, indel = 2e-9),
                    drivers = list(SNV("22", 51240420, "T")))
```

We also need to declare the exposures along the simulation.

```{r}
# add SNV and indel default exposures. This will be used from simulated time 0
# up to the successive exposure change.
m_engine$add_exposure(coefficients = c(SBS13 = 0.2, SBS1 = 0.8))
m_engine$add_exposure(c(ID2 = 0.6, ID13 = 0.2, ID21 = 0.2))

# add a new SNV exposure that will be used from simulated
# time 100 up to the successive exposure change.
m_engine$add_exposure(time = 100, c(SBS5 = 0.3, SBS2 = 0.2, SBS3 = 0.5))
m_engine$add_exposure(time = 120, c(SBS5 = 0.3, SBS2 = 0.2, SBS3 = 0.5,
                                    ID1 = 0.8, ID9 = 0.2))

m_engine
```

We are now ready to build the phylogenetic forest.

```{r}
# place mutations on the sample forest assuming 1000 pre-neoplastic SNVs and
# 500 indels
phylo_forest <- m_engine$place_mutations(sample_forest, 1000, 500)
```

The phylogenetic forest maintains the genome mutations (SBSs, indels, and CNAs)
of all the sampled cells.

```{r}
mutations <- phylo_forest$get_sampled_cell_mutations()

head(mutations)

CNAs <- phylo_forest$get_sampled_cell_CNAs()

head(CNAs)
```

## Sequencing level

We can simulate the sequencing of the collected samples by using the
phylogenetic forest. For each SBS and indel mutation in the phylogenetic
forest, the output reports the number of occurrences in the simulated reads,
the coverage of each mutation locus, and the corresponding VAF in each of the
samples.

```{r}
# simulate the sequencing without normal sample and avoid progress bar
seq_results <- simulate_seq(phylo_forest, coverage = 30,
                            with_normal_sample = FALSE, quiet = TRUE)

head(seq_results$mutations)
```

ProCESS allows to visualise the sequencing results.

```{r, fig.width=4, fig.height=7, fig.alt=c("The VAF histogram labelled by mutation class.")}
plot_VAF_histogram(seq_results, labels = seq_results$mutations["classes"],
                   cuts = c(0.02, 1))
```

Marginal distributions can also be plotted.

```{r, fig.width=4, fig.height=4, fig.alt=c("The VAF marginal distribution S1 vs S2 labelled by mutation class.", "The VAF marginal distribution S1 vs S3 labelled by mutation class.", "The VAF marginal distribution S2 vs S3 labelled by mutation class.")}
plot_VAF_marginals(seq_results, labels = seq_results$mutations["classes"],
                   samples = c("S1", "S2", "S3"))
```

In S1 vs S3 figure, we can some spot passenger mutations occurring in both S1
and S3. Let us identify these mutations.

```{r}
seq_results$mutations %>% filter(classes == "passenger" &
                                   S1.VAF > 0 & S3.VAF > 0)
```
