---
title: "Infinite Site Model"
output: rmarkdown::html_vignette
description: >
  How to enable/disable the infinite site model
vignette: >
  %\VignetteIndexEntry{Infinite Site Model}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options:
  markdown:
    wrap: 72
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

> *Disclaimer:* ProCESS/CLONES implements probability distributions using the
> C++11 random number distribution classes. Since the standard does not specify
> the underlying algorithms, their implementations are left to the compiler.
> Consequently, the simulation output depends on the compiler used to build
> [CLONES](https://github.com/albertocasagrande/CLONES), and the results
> reported in this article may differ from those obtained by the reader.

The mutation engine places mutations on the sampled cell genome according to the
[infinite sites model](https://en.wikipedia.org/wiki/Infinite_sites_model) by
default. In particular, any new mutation is placed on a locus whose context is
mutation-free.

Let us consider the phylogenetic forest as built in
[this article](mutations.md) and verify
whether it satisfies the infinite site model.


```{r, echo=FALSE,  results='hide', message=FALSE}
library("ProCESS")

# set the seed of the random number generator
set.seed(0)

# create a simulation with epigenetic states
sim <- TissueSimulation(epigenetic_states = c("E1", "E2"))

sim$death_activation_level <- 20

# add mutant "A" and set its species rates
sim$add_mutant("A", list("E1" = list(duplication = 0.1, death = 0.08, 
                                     E2 = 0.01),
                         "E2" = list(duplication = 0.1, death = 0.01,
                                     E1 = 0.01)))

# place a cell of A[E1] in the tissue
sim$place_cell("A[E1]", 500, 500)

# let the simulation evolve until "A[E1]" consists of 1000 cells
sim$run_up_to_size("A[E1]", 1000)

bbox_width <- 50
sim$sample_cells("S_1_1",
                 bottom_left = c(480, 480),
                 top_right = c(480 + bbox_width, 480 + bbox_width))

sim$sample_cells("S_1_2",
                 bottom_left = c(500, 500),
                 top_right = c(500 + bbox_width, 500 + bbox_width))

sim$run_up_to_time(sim$get_clock() + 15)

# choose a cell in "A"
cell <- sim$choose_border_cell_in("A")

# add the mutant "B"
sim$add_mutant("B", list("E1" = list(duplication = 0.8, death = 0.05,
                                     E2 = 0.05),
                         "E2" = list(duplication = 0.3, death = 0.05,
                                     E1 = 0.1)))

# let `cell` generate a cell in "B"
sim$mutate_progeny(cell, "B")

# let the simulation evolve until "B[E1]" consists of 7000 cells
sim$run_up_to_size("B[E1]", 7000)

n_w <- n_h <- 25
ncells <- 0.9 * n_w * n_h

bbox <- sim$search_sample(c("A" = ncells), n_w, n_h)
sim$sample_cells("S_2_1", bbox$lower_corner, bbox$upper_corner)

bbox <- sim$search_sample(c("B" = ncells), n_w, n_h)
sim$sample_cells("S_2_2", bbox$lower_corner, bbox$upper_corner)

forest <- sim$get_sample_forest()

# save the sample forest in the file "sample_forest.sff"
forest$save("sample_forest.sff")

# building a mutation engine by using the "demo" setup
m_engine <- MutationEngine(setup_code = "demo")

# add the mutant "A" characterized by one driver SNV on the allele 1 of
# chromosome 22, one indel deletion on the same chromosome, and two CNAs: a
# deletion on the allele 1 of chromosome 22 and an amplification on a
# random allele of the same chromosome. The mutant has two epigenetic states
# and its species "A[E1]" and "A[E2]" have passenger SNV rates 1e-9 and 3e-8,
# respectively, passenger indel rates 1e-9 and 1e-11, respectively, and 
# passenger CNA rates 0 and 1e-11, respectively.
m_engine$add_mutant(mutant_name = "A",
                    passenger_rates = list("E1" = c(SNV = 1e-9, indel = 1e-9),
                                           "E2" = c(SNV = 3e-8, CNA = 1e-11)),
                    drivers = list(SNV("22", 10510210, "C", allele = 1),
                                   Mutation("22", 16085675, "GCCTCCCGA", "G"),
                                   CNA(type = "A", chr = "22",
                                       from = 10303470, len = 200000),
                                   CNA("D", "22", 5010000, 200000,
                                       allele = 1)))

# add the mutant "B" characterized by two driver SNVs on chromosome 22 (no
# CNA) and two epigenetic states. The first SNV is "DGCR8 P26L" and it must
# lay in the allele 1. The remaining SNV is specified by using the SNV function
# as done above. The species "B[E1]" and "B[E2]" have passenger SNV rates 1e-8
# and 5e-8, respectively, and CNA rates 0 for both species.
m_engine$add_mutant("B", list("E1" = c(SNV = 1e-8), "E2" = c(SNV = 5e-8)),
                    list(list("DGCR8 P26L", allele = 1),   # the first SNV
                         SNV("22", 12028576, "G")))       # the second SNV


# add SNV and indel default exposures. This will be used from simulated time 0
# up to the successive exposure change.
m_engine$add_exposure(coefficients = c(SBS13 = 0.2, SBS1 = 0.8))
m_engine$add_exposure(c(ID2 = 0.6, ID13 = 0.2, ID21 = 0.2))

# add a new SNV exposure that will be used from simulated
# time 100 up to the successive exposure change.
m_engine$add_exposure(time = 100, c(SBS5 = 0.3, SBS2 = 0.2, SBS3 = 0.5))

m_engine$add_exposure(time = 120, c(SBS5 = 0.3, SBS2 = 0.2, SBS3 = 0.5,
                                    ID1 = 0.8, ID9 = 0.2))

subjects <- m_engine$get_germline_subjects()

m_engine$set_germline_subject(subjects[2, "sample"])

sample_forest <- load_sample_forest("sample_forest.sff")

# place mutations on the sample forest assuming 1000 pre-neoplastic SNVs and
# 500 indels
phylo_forest <- m_engine$place_mutations(sample_forest, 1000, 500)
```

```{r}
library(dplyr)

# this function verifies whether any mutation arose in
# two unrelated cells represented in a phylogenetic forest
test_infinite_sites_model <- function(phylo_forest) {
  # extract non-germinal SNVs that appear multiple times either
  # in the same cell or different cells
  snvs <- phylo_forest$get_sampled_cell_mutations() %>%
    filter(class != "germinal", .data$type == "SNV") %>%
    count(.data$chr, .data$from, .data$ref, .data$alt) %>%
    filter(n > 1)

  # search for an SNV that independently occurred in two unrelated cells
  first_occurrences <- c()
  row <- 1
  while (length(first_occurrences) < 2 && row <= nrow(snvs)) {
    snv <- SNV(snvs[row, "chr"], snvs[row, "from"],
               ref = snvs[row, "ref"], alt = snvs[row, "alt"])

    first_occurrences <- phylo_forest$get_first_occurrences(snv)
    row <- row + 1
  }

  # if the last handled SNV independently occurred in two unrelated
  # cells at least
  if (length(first_occurrences) >= 2) {

    # print a message containing the two cells
    paste0("SNV('", snv$get_chromosome(), "'',",
           snv$get_position_in_chromosome(), ",'", snv$get_ref(),
           "','", snv$get_alt(),
           "') independently arises in cells ", first_occurrences[1],
           " and ", first_occurrences[2])
  } else {
    print("Every mutation arises exclusively in one cell")
  }
}

# test whether the infinite sites conditions hold in the built forest
test_infinite_sites_model(phylo_forest)
```

This behaviour can be changed by using the mutation engine property
<code>[MutationEngine$infinite_sites_model](../reference/MutationEngine-cash-infinite_sites_model.md)</code>.
This property is a Boolean flag that enables/disables the infinite sites
model.

```{r}
# establish whether the infinite sites model is used
m_engine$infinite_sites_model

# disable it
m_engine$infinite_sites_model <- FALSE
```

When the infinite sites model is disabled,
<code>[MutationEngine$place_mutations()](../reference/MutationEngine-cash-place_mutations.md)</code> may place two mutations
in the same locus of different alleles of the same genome
or the same mutation in the same locus of one allele
of the genomes of two cells that are not each other ancestors.

```{r}
# test whether the infinite sites model is enable
m_engine$infinite_sites_model

# place the mutations on the same sample forest used above
phylo_forest2 <- m_engine$place_mutations(sample_forest, 1000, 500)

# test whether the infinite sites conditions hold in the new forest
test_infinite_sites_model(phylo_forest2)
```

