---
title: "Simulating Mutations"
output: rmarkdown::html_vignette
description: >
  How to simulate mutations on sampled cells
vignette: >
  %\VignetteIndexEntry{Simulating Mutations}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options:
  markdown:
    wrap: 72
bibliography: references.bib
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

> *Disclaimer:* ProCESS/CLONES implements probability distributions using the
> C++11 random number distribution classes. Since the standard does not specify
> the underlying algorithms, their implementations are left to the compiler.
> Consequently, the simulation output depends on the compiler used to build
> [CLONES](https://github.com/albertocasagrande/CLONES), and the results
> reported in this article may differ from those obtained by the reader.

ProCESS/CLONES can simulate genomic mutations in the cells represented in
a `SampleForest` according to the specified SBS and indel mutational
signatures (see [@alexandrov2020repertoire]).
This process is performed by the class `MutationEngine`, which also takes
into account the mutation rate of the simulated species and gives the
chance to dynamically change the signatures.

### Setting Up Mutation Engine

The creation of an object of the type `MutationEngine` requires downloading
the reference sequence and the signature files and building the
corresponding two signature indices. The function `MutationEngine()`
performs all these steps in a single call.

This function is quite flexible and allows customising the mutation engine in
many ways. However, the vast majority of ProCESS users would aim for a standard
setup, for instance, involving the human genome.

Because of this, ProCESS provides some predefined setups.
A list containing them can be obtained by invoking the function
`get_mutation_engine_codes()`.

```{r}
library("ProCESS")

get_mutation_engine_codes()
```

Let us build a `MutationEngine` by using the `"demo"` setup.

```{r, echo=FALSE,  results='hide', message=FALSE}
unlink("demo", recursive = TRUE)
```

```{r}
# building a mutation engine by using the "demo" setup
m_engine <- MutationEngine(setup_code = "demo")

m_engine
```

The above call creates the directory `demo`,
downloads all the data required by the mutation
engine, and builds it.

```{r}
dir.exists("demo")

list.files("demo")
```

The execution of the function `MutationEngine()` may take some time, but
it is meant to be performed one for all and, as long as the user does not need
to change the reference genome or the signature files, it is no longer
required. In this spirit, any call to `MutationEngine()` exclusively performs
the building process sub-tasks that have not been already fulfilled.

```{r}
# building a mutation engine by using the "demo" set-up configuration
m_engine <- MutationEngine(setup_code = "demo")
```

#### Mutant Genetic Characterization

Once the mutation engine has been built, we can define a mutant genotype and
declare species mutation rates.

Let us consider the simulation of Section "_Two populations with
epigenetic state_" in [this article](sampling.md). It involves the two
mutants `A` and `B`. Both of them have two possible epigenetic states,
`E1` and `E2`, leading to the four species `A[E1]`, `A[E2]`, `B[E1]`, and 
`B[E2]`, respectively.
Each of these species has a passenger mutation rate that must be
declared to the mutation engine before the sample forest is labelled.
Let 1e-9, 3e-8, 8e-7, and 5e-8 be the passenger SNV rates for the
species `A[E1]`, `A[E2]`, `B[E1]`, and `B[E2]`, respectively. Let the indel
rates be 0 for all the species, but `A[E1]`, which has an indel rate of 1e-8.
Furthermore, let 0, 1e-11, 0, and 0 be the passenger CNA rates of the same
species, respectively.

The two mutants may also be genetically characterised by some
driver mutations.
The driver mutations associated with each mutant must occur in any cell
belonging to that mutant. Hence, they must be
declared to the mutant engine before the labelling.

The method <code>[MutationEngine$add_mutant()](../reference/MutationEngine-cash-add_mutant.md)</code>
takes care of all these declarations.

For the sake of example, let us assume that `A` is characterised by one
driver mutation on chromosome 22, while `B` has three driver mutations on
the same chromosome.

The genetic specification of the mutant `A` can be declared as it follows.

```{r, echo=FALSE,  results='hide', message=FALSE}
m_engine <- MutationEngine(setup_code = "demo")
```

```{r}
# add the mutant "A" characterized by one driver SNV on the allele 1 of
# chromosome 22, one indel deletion on the same chromosome, two CNAs (a
# deletion on the allele 1 of chromosome 22 and an amplification on a random
# allele of the same chromosome) and, finally, a whole genome doubling event
# (WGD). The mutant has two epigenetic states and its species "A[E1]" and
# "A[E2]" have passenger SNV rates 1e-9 and 3e-9, respectively, passenger indel
# rates 1e-9 and 0, respectively, and passenger CNA rates 0 and 1e-11,
# respectively.
m_engine$add_mutant(mutant_name = "A",
                    passenger_rates = list("E1" = c(SNV = 1e-9, indel = 1e-9),
                                           "E2" = c(SNV = 3e-9, CNA = 1e-11)),
                    drivers = list(SNV("22", 46510210, "C", allele = 1),
                                   Mutation("22", 16085675, "GCCTCCCGA", "G"),
                                   CNA(type = "A", chr = "22",
                                       from = 10303470, len = 200000),
                                   CNA("D", "22", 5010000, 200000,
                                       allele = 1),
                                   WGD))

m_engine
```

In the above code, the driver CNAs, SNVs, and indels are declared by calling
the functions `CNA()`, `SNV()`, and `Mutation()`, respectively. These functions
allow specifying the allele in which the mutation must lie.
The mutant `A` is also characterised by _whole genome doubling_ (`WGD`):
a genomic event that simultaneously duplicates all the chromosome alleles.

The driver mutations are applied in the order specified.
For instance, the following two pieces of code specify two different
genomic characterisations for the same mutant `E`.
```{r, eval=FALSE}
m_engine$add_mutant("E", passenger_rates,
                    drivers = list(SNV("22", 10510210, "C", allele = 1),
                                   WGD))
```

```{r, eval=FALSE}
m_engine$add_mutant("E", passenger_rates,
                    drivers = list(WGD,
                                   SNV("22", 10510210, "C", allele = 1)))
```

The former snippet places an SNV and, afterwards, duplicates all the alleles
producing a genome in which two alleles contain the placed SNV. Instead, the
latter snippet requires a whole genome doubling event and, then, places the
SNV yielding a single occurrence of the SNV in the final genome.

As far as SNVs and indels are concerned, ProCESS provides users with a more
compact and, at times, convenient approach. The mutation engine stores a list
of known driver mutations and labels each with a code that can be used during
mutant declaration. The corresponding data frame can be obtained
by using the method
<code>[MutationEngine$get_known_drivers()](../reference/MutationEngine-cash-get_known_drivers.md)</code>.

```{r}
library(dplyr)

m_engine$get_known_drivers() %>% filter(chr == "22") %>% head()
```

The code of the known driver mutations can be used in place of the full
specification as follows.

```{r}
# add the mutant "B" characterized by two driver SNVs on chromosome 22 (no
# CNA) and two epigenetic states. The first SNV is "DGCR8 P26L" and it must
# lay in the allele 1. The remaining SNV is specified by using the SNV function
# as done above. The species "B[E1]" and "B[E2]" have passenger SNV rates 8e-9
# and 2e-8, respectively, and CNA rates 0 for both species.
m_engine$add_mutant("B", list("E1" = c(SNV = 8e-9), "E2" = c(SNV = 2e-8)),
                    list(list("DGCR8 P26L", allele = 1),  # the first SNV
                         SNV("22", 12028576, "G")))       # the second SNV

m_engine
```

#### Mutational Exposures

The probability of a mutation occurring depends on both its genomic and
environmental context.

A _signature_ is a mutation probability distribution over mutation contexts or
mutation structure. _SBS (single-base substitution) signatures_ provide
for any genomic context (i.e., a triplet of bases) the probability that an SNV
occurs on that context. On the contrary, _ID (indel) signatures_ associate
the probability of an indel to its length and structure
(see [@alexandrov2020repertoire]).

The signature depends on the environmental context and, as a result, more
than one signature may be active at the same time with different probabilities.
A _mutational exposure_ (or _exposure_) is a discrete probability distribution
among signatures.

To simulate passenger mutations of a given type, we need to specify a
default exposure for that type. This can be achieved as follows.

```{r}
# add SNV and indel default exposures. This will be used from simulated time 0
# up to the successive exposure change.
m_engine$add_exposure(coefficients = c(SBS13 = 0.2, SBS1 = 0.8))
m_engine$add_exposure(c(ID2 = 0.6, ID13 = 0.2, ID21 = 0.2))
```

Further exposures can also be defined by specifying an activation time
for each of them, i.e., the new exposures will be used
from that until the next exposure change.

```{r}
# add a new SNV exposure that will be used from simulated
# time 100 up to the successive exposure change.
m_engine$add_exposure(time = 100, c(SBS5 = 0.3, SBS2 = 0.2, SBS3 = 0.5))

m_engine
```

The default exposures and the simultaneous changes in SNV and ID exposures
can be specified by a single call to the function
<code>[MutationEngine$add_exposure()](../reference/MutationEngine-cash-add_exposure.md)</code> as follows.

```{r}
m_engine$add_exposure(time = 120, c(SBS5 = 0.3, SBS2 = 0.2, SBS3 = 0.5,
                                    ID1 = 0.8, ID9 = 0.2))

m_engine
```

#### Passenger CNAs and Tumour Type

The passenger CNAs applied during the simulation depend on tumour type.
The type of tumour can be specified when building the mutation engine
by using the parameters `tumour_type` or `tumour_study`. For instance,
if the passenger CNA file used to build the mutation engine contains
some of the data identified in breast carcinoma in the UK,
the specific CNA can be applied during the simulation as follows.

```{r, eval=FALSE}
m_engine <- MutationEngine(setup_code = "demo",
                           tumour_type = "BRCA",
                           tumour_study = "UK")
```

The complete list of available tumour types in a set-up
can be obtained by using the function `get_available_tumours_in()`.

```{r}
get_available_tumours_in(setup_code = "demo") %>% head()
```

#### Germline Mutations

ProCESS allows users to apply germline mutations from one of the subjects in
the germline data used to build the mutation engine. This feature will enable
users to simulate a specific cancer type’s evolution on an individual with the
desired gender and ethnicity.

The available subjects, together with their sex and ethnicity, can
be obtained by using the method
<code>[MutationEngine$get_germline_subjects()](../reference/MutationEngine-cash-get_germline_subjects.md)</code>.

```{r}
subjects <- m_engine$get_germline_subjects()

subjects
```

The column `sample` contains the names of the available subjects.
The columns `pop` and `super_pop` report the subjects' population
and super-population codes. The last column,
`gender`, includes the subject gender.

The method
<code>[MutationEngine$get_population_descriptions()](../reference/MutationEngine-cash-get_population_descriptions.md)</code>
clarifies the meaning of the codes reported in the pop columns.

```{r}
m_engine$get_population_descriptions()
```

The method
<code>[MutationEngine$get_active_germline()](../reference/MutationEngine-cash-get_active_germline.md)</code> returns the
the active germline subject.

```{r}
m_engine$get_active_germline()
```

Users can change the germline subject using the method
<code>[MutationEngine$set_germline_subject()](../reference/MutationEngine-cash-set_germline_subject.md)</code>.

When a subject is selected for the first time, ProCESS builds
a binary representation of the subject's genome and save
it for future use. This step may take a few minutes. However,
all the successive selections of the same subject directly
load the binary file.

```{r}
m_engine$set_germline_subject(subjects[2, "sample"])

m_engine$get_active_germline()
```

### Building Phylogenetic Forests

The configured mutation engine can be used to label each node in a
sample forest with mutations.

Since the mutation engine has been configured to deal with the
simulation performed in Section "_Two populations with
epigenetic state_" in [this article](sampling.md),
we can load its sample forest from the file `"sample_forest.sff"` as saved
there.

```{r}
sample_forest <- load_sample_forest("sample_forest.sff")

# place mutations on the sample forest assuming 1000 pre-neoplastic SNVs and
# 500 indels
phylo_forest <- m_engine$place_mutations(sample_forest, 1000, 500)

phylo_forest
```

The phylogenetic forest stores all mutations, labelling the sampled cells
represented by its leaves. Users can retrieve such data by using the methods
<code>[PhylogeneticForest$get_sampled_cell_mutations()](../reference/PhylogeneticForest-cash-get_sampled_cell_mutations.md)</code>
and
<code>[PhylogeneticForest$get_sampled_cell_CNAs()](../reference/PhylogeneticForest-cash-get_sampled_cell_CNAs.md)</code>.

```{r}
library(dplyr)

# select the first mutations among all the mutations occurring in
# the genomes of the sampled cells
phylo_forest$get_sampled_cell_mutations() %>% head()

# select the first CNAs among all the mutations occurring in
# the genomes of the sampled cells
phylo_forest$get_sampled_cell_CNAs() %>% head()

# get the sampled cells
sampled_cells <- phylo_forest$get_nodes() %>%
  filter(!is.na(.data$sample))

# show the first of them
sampled_cells %>% head()

# get the identifier of the 3rd cell in `sampled_cells`
cell_id <- sampled_cells[3, 1]

# get the SNVs and the indels of the 3rd cell in `sampled_cells`
phylo_forest$get_cell_mutations(cell_id) %>% head()

# get the CNAs of the 3rd cell in `sampled_cells`
phylo_forest$get_cell_CNAs(cell_id) %>% head()
```

The method
<code>[PhylogeneticForest$get_samples_info()](../reference/PhylogeneticForest-cash-get_samples_info.md)</code> is analogous to
<code>[SampleForest$get_samples_info()](../reference/SampleForest-cash-get_samples_info.md)</code>: it returns a data frame containing
information about the forest samples. However, the former method adds two
columns to the result of the latter: `DNA_quantity` and
`equivalent_normal_cells`. The former column reports the total quantity
of tumoral DNA in the sample, i.e., the sum of the lengths of all
alleles in the sample. This quantity is a natural number,
nevertheless, it is stored as a real number as it usually exceeds
the largest natural number that can be natively represented by R.
The column "`equivalent_normal_cells`" instead contains the number of
normal cells that contain as much DNA as the sample tumour cells.

```{r}
phylo_forest$get_samples_info()
```

The method <code>[PhylogeneticForest$get_germline_mutations()](../reference/PhylogeneticForest-cash-get_germline_mutations.md)</code> returns
the SNVs and the indels in the germline.

```{r}
# extract the germline mutation
phylo_forest$get_germline_mutations() %>% head()
```

Users can also identify the cell in which a mutation emerged
even when the cell was not sampled.

```{r}
# select one of the mutations
mutation_row <- phylo_forest$get_cell_mutations(cell_id)[2, ]

# rebuild the corresponding mutation
mutation <- Mutation(mutation_row["chr"][1, ],
                     mutation_row["from"][1, ],
                     mutation_row["ref"][1, ],
                     mutation_row["alt"][1, ])

# get the identifier of the oldest cells in which the mutation occurs
phylo_forest$get_first_occurrences(mutation)
```

The exposures used in placing the mutations on the cells in the
phylogenetic forest can be obtained by using the method
<code>[PhylogeneticForest$get_exposures()](../reference/PhylogeneticForest-cash-get_exposures.md)</code>.

```{r}
# get the exposures used in placing the mutations
phylo_forest$get_exposures()
```

The method <code>[MutationEngine$get_bulk_allelic_fragmentation()](../reference/MutationEngine-cash-get_bulk_allelic_fragmentation.md)</code> 
returns a data frame
reporting the allelic type per genome fragment.

```{r}
# get the name of the first sample
sample_name <- phylo_forest$get_samples_info()[["name"]][1]

# print the bulk allelic fragmentation
phylo_forest$get_bulk_allelic_fragmentation(sample_name) %>% head()
```

Instead, the method <code>[MutationEngine$get_cell_allelic_fragmentation()](../reference/MutationEngine-cash-get_cell_allelic_fragmentation.md)</code> 
returns the allelic fragmentation per cell.

```{r}
# print the cell allelic fragmentation
phylo_forest$get_cell_allelic_fragmentation() %>% head()
```

The details about the SNV and indel signatures adopted during the evolution are
available in the mutation engine and they can be retrieved by using the methods
<code>[MutationEngine$get_SNV_signatures()](../reference/MutationEngine-cash-get_SNV_signatures.md)</code> 
and
<code>[MutationEngine$get_indel_signatures()](../reference/MutationEngine-cash-get_indel_signatures.md)</code>.

```{r}
# get the SNV signatures used in placing the mutations
m_engine$get_SNV_signatures()[1:6, 1:5]

# get the indel signatures used in placing the mutations
m_engine$get_indel_signatures()[1:6, 1:5]
```

Finally, the data of the subject whose germline corresponds to
wild-type genome in the phylogenetic forest can be
obtained by the method
<code>[PhylogeneticForest$get_germline_subject()](../reference/PhylogeneticForest-cash-get_germline_subject.md)</code>.

```{r}
phylo_forest$get_germline_subject()
```


### Storing Phylogenetic Forests

As in the case of the sample forests, the phylogenetic forests
can be saved by using the method 
<code>[PhylogeneticForest$save()](../reference/PhylogeneticForest-cash-save.md)</code> 
and load by the function `load_phylogenetic_forest()`.

```{r}
# save the phylogenetic forest in the file "phylo_forest.sff"
phylo_forest$save("phylo_forest.sff")

# loading the saved forest
loaded_phylo_forest <- load_phylogenetic_forest("phylo_forest.sff")

loaded_phylo_forest
```

### Getting and Setting the Reference Genome Path

The phylogenetic forest object contains the reference genome FASTA file path.
The methods 
<code>[PhylogeneticForest$get_reference_path()](../reference/PhylogeneticForest-cash-get_reference_path.md)</code> and
<code>[PhylogeneticForest$set_reference_path()](../reference/PhylogeneticForest-cash-set_reference_path.md)</code> 
can be used to get the path and set it, respectively.
```{r}
phylo_forest$get_reference_path()

phylo_forest$set_reference_path("demo/reference.fasta")

phylo_forest$get_reference_path()
```
