---
title: "Iterating over Forest Nodes"
output: rmarkdown::html_vignette
description: >
  How to iterate over forest node
vignette: >
  %\VignetteIndexEntry{Iterating over Forest Nodes}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options:
  markdown:
    wrap: 72
bibliography: references.bib
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

> *Disclaimer:* ProCESS/CLONES implements probability distributions using the
> C++11 random number distribution classes. Since the standard does not specify
> the underlying algorithms, their implementations are left to the compiler.
> Consequently, the simulation output depends on the compiler used to build
> [CLONES](https://github.com/albertocasagrande/CLONES), and the results
> reported in this article may differ from those obtained by the reader.

ProCESS provides a practical interface to iterate
over the nodes of any forest by using the natures
<code>[SampleForestNodeTour](../reference/SampleForestNodeTour.md)</code>
and
<code>[PhylogeneticForestNodeTour](../reference/PhylogeneticForestNodeTour.md)</code>.

Let us reload the sample forest produced in
[this article](sampling.md).

```{r}
library("ProCESS")

# load the phylogenetic forest produced by sampling.md
sample_forest <- load_forest("sample_forest.sff")

# show some information about the forest
sample_forest
```

The function `get_node_tour()` accepts as parameter a forest and
returns either a
<code>[SampleForestNodeTour](../reference/SampleForestNodeTour.md)</code>
or a
<code>[PhylogeneticForestNodeTour](../reference/PhylogeneticForestNodeTour.md)</code>
object depending on the forest type.
These natures implement iterators over the forest nodes and, at each
instant, point to one of the forest nodes.
Their objects provide the properties `node` and `done`, as well as the method
`step()`. The property `node` maintains the node pointed by the iterator and
is either a
<code>[SampleForestNode](../reference/SampleForestNodeTour.md)</code>
or a
<code>[PhylogeneticForestNode](../reference/PhylogeneticForestNodeTour.md)</code>
object depending on the type of the forest passed to `get_node_tour()`.
Instead, the tour property `done` is a
Boolean value that is `TRUE` if and only if the end of the iteration
has been reached. Finally, the method `step()` lets the iterator move to
the next node when available.

```{r}
node_tour <- get_node_tour(sample_forest)
node_tour

# get the first node in the tour
node_tour$node

# test whether the tour has been completed
node_tour$done

# go to the next node
node_tour$step()

# the second value of the labelling tour
node_tour$node
```

Since the forest type is
<code>[SampleForest](../reference/SampleForest.md)</code>,
`get_node_tour()` returns a
<code>[SampleForestNodeTour](../reference/SampleForestNodeTour.md)</code>
object and the property `node` has type
<code>[SampleForestNode](../reference/SampleForestNode.md)</code>.

Let us reload the phylogenetic forest produced in
[this article](mutations.md) and build a new tour.

```{r}
# load the phylogenetic forest produced by mutations.md
phylo_forest <- load_forest("phylo_forest.pff")

# show some information about the forest
phylo_forest

# build a tour over `phylo_forest` nodes
node_tour <- get_node_tour(phylo_forest)
node_tour

# get the first node in the tour
node_tour$node
```

The tour visits nodes in depth first search order. Thus, the
first node to be visited is a forest root.

```{r}
# check whether the node is a forest root
node_tour$node$is_root

# check whether the node is a forest leaf
node_tour$node$is_leaf

# get the name of the sample that collected the cell corresponding to
# the node
node_tour$node$sample_name
```

#### Visiting the Forest Leaves

By setting the optional Boolean parameter `only_leaves` to `TRUE`,
`get_node_tour()` produces an iterator over the forest leaves.

```{r}
# build a tour over `phylo_forest` node
node_tour <- get_node_tour(phylo_forest, only_leaves = TRUE)

# check whether the first node is a forest leaf
node_tour$node$is_leaf

# move to the next leaf
node_tour$step()

# check whether the second node is a forest leaf
node_tour$node$is_leaf

# get the name of the sample that collected the cell corresponding to
# the node
node_tour$node$sample_name
```

#### Labelling Nodes

The function `get_node_tour()` accepts as an optional parameter
`labelling_functor` and generates the node labels by applying the labelling
functor to the forest nodes in a depth-first visit order.

The labelling functor assigns a label to each node within the forest and must
have type `function(parent_label, node)` where the type of `node` is either a
<code>[SampleForestNode](../reference/SampleForestNode.md)</code> object or a
<code>[PhylogeneticForestNode](../reference/PhylogeneticForestNode.md)</code>
object depending on the forest type.

A labelling function can label each node with the identifier of the
corresponding cell.
```{r}
# a function that label each node by the identifier of the corresponding cell
id_labelling <- function(label, node) {
  node$cell_id
}

# a function that label each node by a list of the identifier and species
# of the corresponding cell
id_species_labelling <- function(parent_label, node) {
  list(cell_id = node$cell_id, species = node$species_name)
}

# a function that label each node by the list containing its cell
# identifier and the identifier of its ancestors
ancestor_species_labelling <- function(parent_label, node) {
  # when the node a root its parent label is NULL
  if (is.null(parent_label)) {
    list(node$cell_id)
  } else {
    unique(append(parent_label, node$species_name))
  }
}
```

The labelling function can also access to external variables. For instance,
we can label forest nodes by using values sampled in a list by using the
following labelling function.

```{r}
values <- 1:100

# this labelling function labels each node by randomly sampling the
# content of values
random_labelling <- function(parent_label, node) {
  sample(values, 1)
}
```

When the optional parameter `labelling_functor` is provided, the label of
the pointed forest node is available in the tour additional property
`label`.

```{r}
# generate a node tour labelling each node by the identifiers
# of the corresponding cells
node_tour <- get_node_tour(phylo_forest, id_labelling)

# the first node of the tour
node_tour$node

# the label of the first node
node_tour$label

# go to the next node
node_tour$step()

# the label of the second node
node_tour$label

# generate a node tour labelling each node by the ids of its ancestors
node_tour <- get_node_tour(phylo_forest, ancestor_species_labelling)

# go to the first node whose associated cell belongs to "B[E1]"
while (node_tour$node$species_name != "B[E1]") {
  node_tour$step()
}

node_tour$node

# show the first five ancestors of the node
node_tour$label[1:5]
```

#### Producing the Simulated Genomes

Behind the labels produced by the labelling function,
<code>[PhylogeneticForestNodeTour](../reference/PhylogeneticForestNodeTour.md)</code>
can also generate the genomes of the forest nodes. This is achieved by using the
`get_node_tour()`'s optional Boolean parameter `with_genomes`.
When `with_genomes` is set to `TRUE`, the tour additional property `genome` contains a
<code>[GenomeMutations](../reference/GenomeMutations.md)</code> object
representing the genome of the cell modelled by the pointed node.

```{r}
# generate a node tour with genomes
node_tour <- get_node_tour(phylo_forest, with_genomes = TRUE)

node_tour$genome
```

Notice that the genome of each node can by generated by using the
<code>[PhylogeneticForestNode](../reference/PhylogeneticForestNode.md)</code>
method <code>[get_genome()](../reference/PhylogeneticForestNode.md)</code>.
However, the asymptotic time cost of this method is linear in the number of
ancestors of the node, while the overall cost of a tour is linear in the
number of nodes. Hence, iterating over the the node/leaf genomes is much
faster than iterating over the nodes and generating their genomes by using
<code>[get_genome()](../reference/PhylogeneticForestNode.md)</code>.

##### Accessing Cell Genome Data

The class <code>[GenomeMutations](../reference/GenomeMutations.md)</code>
provides methods to access cell genome data.

The method
<code>[GenomeMutations$get_allele_fragments()](../reference/GenomeMutations-cash-get_allele_fragments.md)</code>
returns a data frame containing all the allele fragments in cell genome.

```{r}
# get the genome of the pointed node
genome <- node_tour$genome

# see the genome structure (i.e., chromosomes, alleles, allele fragments).
# Here we exclusively see chromosome 22 because of the mutation engine
# demo setup.
genome$get_allele_fragments()
```

The mutations accumulated in the cell genome can be obtained calling
the method <code>[GenomeMutations$get_mutations()](../reference/GenomeMutations-cash-get_mutations.md)</code>.

```{r}
# get the first mutations in the genome
head(genome$get_mutations())
```

Users can also extract a fragment of the cell genome. For instance, the
following snippet generates a <code>[GenomeFragment](../reference/GenomeFragment.md)</code> object representing the fragment
beginning at position 16085655, chromosome 22, and allele 1 whose length is 100
nucleotides.

```{r}
# extract a fragment from the genome
fragment <- genome$get_fragment("22", 1, 16085655, 100)

fragment
```

#### Genome Fragment Data

Behind its position, any <code>[GenomeFragment](../reference/GenomeFragment.md)</code>
object maintains the DNA sequence
corresponding to the fragment, the mutations occurring to it, and its CIGAR
code [@li2009sam] with respect to the reference genome.

```{r}
# the region covered in the reference genome
fragment$get_covered_reference_region()

# the fragment DNA sequence
fragment$sequence

# the mutations laying in the fragment
fragment$get_mutations()

# the CIGAR code for the fragment
fragment$get_CIGAR()
```