---
title: "Forest Node Labelling"
output: rmarkdown::html_vignette
description: >
  How to label forest nodes
vignette: >
  %\VignetteIndexEntry{Forest Node Labelling}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options:
  markdown:
    wrap: 72
bibliography: references.bib
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

> *Disclaimer:* ProCESS/CLONES implements probability distributions using the
> C++11 random number distribution classes. Since the standard does not specify
> the underlying algorithms, their implementations are left to the compiler.
> Consequently, the simulation output depends on the compiler used to build
> [CLONES](https://github.com/albertocasagrande/CLONES), and the results
> reported in this article may differ from those obtained by the reader.

ProCESS supports recursive labelling of forest nodes. The function
`get_label_tour()` accepts a forest and a labelling function as input and
generates an iterator over the forest node labels by applying the labelling
function to the forest nodes in a depth-first visit order. The labelling
function takes as input the label of the parent node, if available, and a node,
and computes the label of the node itself. 

#### Labelling Functions

The labelling function assigns a label to each node within the forest and must
have type `function(parent_label, node)` where the type of `node` is either a
<code>[SampleForestNode](../reference/SampleForestNode.md)</code> object or a
<code>[PhylogeneticForestNode](../reference/PhylogeneticForestNode.md)</code>
object depending on the
forest type. These two classes expose properties of cells represented by
forest nodes. For instance, the cell identifiers or the cell species. The
following examples only report some of the available properties. Please, refer
to <code>[SampleForestNode](../reference/SampleForestNode.md)</code> 
and <code>[PhylogeneticForestNode](../reference/PhylogeneticForestNode.md)</code>
for a complete overview.

A labelling function can label each node with the identifier of the
corresponding cell.
```{r}
# a function that label each node by the identifier of the corresponding cell
id_labelling <- function(label, node) {
  node$cell_id
}

# a function that label each node by a list of the identifier and species
# of the corresponding cell
id_species_labelling <- function(parent_label, node) {
  list(cell_id = node$cell_id, species = node$species)
}

# a function that label each node by the list containing its cell
# identifier and the identifier of its ancestors
ancestor_id_labelling <- function(parent_label, node) {
  # when the node a root its parent label is NULL
  if (is.null(parent_label)) {
    list(node$cell_id)
  } else {
    append(parent_label, node$cell_id)
  }
}

# a function to label each node by a list consisting of two Boolean values. The
# first is named `leaf` and is `TRUE` if and only if the node is a leaf of the
# forest. The second is named `root` and is `TRUE` if and only if the node is a
# root.
node_prop_labelling <- function(parent_label, node) {
  list(leaf = node$is_leaf, root = node$is_root)
}
```

The labelling function can also access to external variables. For instance,
we can label forest nodes by using values sampled in a list by using the
following labelling function.

```{r}
values <- 1:100

# this labelling function labels each node by randomly sampling the
# content of values 
random_labelling <- function(parent_label, node) {
  sample(values, 1)
}
```

#### Iterating over Forest Nodes

Let us reload the phylogenetic forest produced in
[this article](mutations.md).

```{r}
library("ProCESS")

# load the phylogenetic forest produced by mutations.md
phylo_forest <- load_phylogenetic_forest("phylo_forest.sff")

# show some information about the forest
phylo_forest
```

Each node of the forest represents a cell that can be labelled by the
function `get_label_tour()`. This function returns either a
<code>[SampleForestLabelTour](../reference/SampleForestLabelTour.md)</code>
or a
<code>[PhylogeneticForestLabelTour](../reference/PhylogeneticForestLabelTour.md)</code>
object depending on the type of the forest parameter.
These classes implement iterators over the forest nodes and, at each
instant, refer to one of the nodes.
They provide the properties `value` and `done`, as well as the method
`step()`. The property `value` is a named list consisting of both the
identifier of the cell corresponding to the currently referred node (name
`cell_id`) and the node label (name `label`). Instead, the property `done`
is a Boolean value that is `TRUE` if and only if the end of the iteration
has been reached. Finally, the method `step()` lets the iterator move to
the next node when available.

```{r}
# generate a label tour labelling each node by the identifiers
# of the corresponding cells
label_tour <- get_label_tour(phylo_forest, id_labelling)

# the first value of the labelling tour
label_tour$value

# test whether the tour has been completed
label_tour$done

# go to the next node
label_tour$step()

# the second value of the labelling tour
label_tour$value

# generate a label tour labelling each node by its properties
label_tour <- get_label_tour(phylo_forest, node_prop_labelling)

# the first value of the labelling tour. Notice that the first
# node is a root, not a leaf
label_tour$value
```

#### Iterating over Forest Leaves

The function `get_label_tour()` has the optional Boolean parameter `only_leaves`
that, when set to `TRUE` allows to iterate over the leaves.

```{r}
# generate a label tour labelling each forest leaf by its properties
label_tour <- get_label_tour(phylo_forest, node_prop_labelling,
                             only_leaves = TRUE)

# since the parameter `only_leaves` is set to `TRUE`, in this case, the first
# node is not a root, but a leaf
label_tour$value

# ... so are all labels in the tour
label_tour$step()

label_tour$value
```

### Producing the Simulated Genomes

Behind the labels produced by the labelling function, 
<code>[PhylogeneticForestLabelTour](../reference/PhylogeneticForestLabelTour.md)</code>
can also label each node by the genome of the
cell corresponding to the referred node. This is achieved by using the
optional Boolean parameter `with_genome` of the function `get_label_tour()`.
When `with_genome` is set to `TRUE`, the property `value` contains a further
element, whose name is `genome`, representing the cell's
genome. In particular, this value is a 
<code>[GenomeMutations](../reference/GenomeMutations.md)</code> object.

```{r}
# generate a label tour labelling each forest leaf by its properties
label_tour <- get_label_tour(phylo_forest, node_prop_labelling,
                             only_leaves = TRUE, with_genomes = TRUE)

label_tour$value
```

#### Accessing Cell Genome Data

The class <code>[GenomeMutations](../reference/GenomeMutations.md)</code> provides methods to access cell genome data.

The method
<code>[GenomeMutations$get_allele_fragments()](../reference/GenomeMutations-cash-get_allele_fragments.md)</code>
returns a data frame containing all the allele fragments in cell genome.

```{r}
# get the genome of the current label tour value
genome <- label_tour$value$genome

# see the genome structure (i.e., chromosomes, alleles, allele fragments).
# Here we exclusively see chromosome 22 because of the mutation engine
# demo setup.
genome$get_allele_fragments()
```

The mutations accumulated in the cell genome can be obtained calling
the method <code>[GenomeMutations$get_mutations()](../reference/GenomeMutations-cash-get_mutations.md)</code>.

```{r}
# get the first mutations in the genome
head(genome$get_mutations())
```

Users can also extract a fragment of the cell genome. For instance, the
following snippet generates a <code>[GenomeFragment](../reference/GenomeFragment.md)</code> object representing the fragment
beginning at position 16085655, chromosome 22, and allele 1 whose length is 100
nucleotides.

```{r}
# extract a fragment from the genome
fragment <- genome$get_fragment("22", 0, 16085655, 100)

fragment
```

#### Genome Fragment Data

Behind its position, any <code>[GenomeFragment](../reference/GenomeFragment.md)</code>
object maintains the DNA sequence
corresponding to the fragment, the mutations occurring to it, and its CIGAR
code [@li2009sam] with respect to the reference genome.

```{r}
# the region covered in the reference genome
fragment$get_covered_reference_region()

# the fragment DNA sequence
fragment$get_sequence()

# the mutations laying in the fragment
fragment$get_mutations()

# the CIGAR code for the fragment
fragment$get_CIGAR()
```

#### Iterating over Genomes without Labels

The function `get_genome_tour()` takes as input a
<code>[PhylogeneticForest](../reference/PhylogeneticForest.md)</code>
object and returns an iterator that produce the simulated cell genomes
without any further node labels. The returned value is a
<code>[PhylogeneticForestLabelTour](../reference/PhylogeneticForestLabelTour.md)</code>
whose property `value` does not have the `label` element.

```{r}
genome_tour <- get_genome_tour(phylo_forest)

genome_tour$value

genome_tour$step()

genome_tour$value
```

As in the case of `get_label_tour()`, the function `get_genome_tour()` as an
optional Boolean parameter `only_leaves` that, when set to `TRUE`, allows
to focus on the genomes of forest leaves.

```{r}
genome_tour <- get_genome_tour(phylo_forest, only_leaves = TRUE)

genome_tour$value

genome_tour$step()

genome_tour$value
```
