---
title: "Formula-based simulation constraints"
output: rmarkdown::html_vignette
description: >
  Simulating a tissue evolution until a simulation state formula
  does not hold
vignette: >
  %\VignetteIndexEntry{Formula-based simulation constraints}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options:
  markdown:
    wrap: 72
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

> *Note:* This article presents advanced topics on tissue simulation. Refer to
> [this article](tissue_simulation.md) for an introduction on the subject.

> *Disclaimer:* ProCESS/CLONES implements probability distributions using the
> C++11 random number distribution classes. Since the standard does not specify
> the underlying algorithms, their implementations are left to the compiler.
> Consequently, the simulation output depends on the compiler used to build
> [CLONES](https://github.com/albertocasagrande/CLONES), and the results
> reported in this article may differ from those obtained by the reader.

ProCESS implements a first-order unquantified logic having variables
representing the cardinality of the species, the number of events
fired in a species (being duplications, deaths, or switches), and
the simulation time. These variables and real values are summed by
`+`, subtracted by `-`, and multiplied by `*` and form expressions.
The expressions are then compared with the standard semantics by
`>`, `>=`, `==`, `!=`, `<=`, and `<` to form relations.
A formula in this language is either a relation, the conjunction
of two formulas (`&`), or the disjunction of two formulas (`|`).

Any formula in the above language expresses a condition on the simulation
status. It can be used as a parameter of the method
<code>[TissueSimulation$run_until()](../reference/TissueSimulation-cash-run_until.md)</code>
to let the simulation evolve until the condition no longer holds.

### Variables

The variables represent one of the following quantities:

-  the cardinality of a species;
-  the number of fired deaths, duplications, or switches in a species;
-  the elapse simulation time.

All the above variables can be built by using the method
<code>[TissueSimulation$var()](../reference/TissueSimulation-cash-var.md)</code>.
When the parameter is the string `"Time"`, the elapsed simulation time
variable is returned.

```{r}
library("ProCESS")

# set the seed of the random number generator
set.seed(0)

# build a tissue simulation with two epigenetic states
sim <- TissueSimulation(epigenetic_states = c("E1", "E2"))

# add the mutant "A"
sim$add_mutant(name = "A", list(E1 = list(duplication = 0.2, death = 0.1,
                                          E2 = 0.01),
                                E2 = list(duplication = 0.08, death = 0.01,
                                          E1 = 0.01)))

# get the variable representing the simulation time
v_time <- sim$var("Time")

v_time
```

When the parameter is the name of a species,
a variable representing the cardinality of the species is built.

```{r}
# get the variable representing the cardinality of A[E1] in sim
va1 <- sim$var("A[E1]")
va1

# get the variable representing the cardinality of A[E2] in sim
va2 <- sim$var("A[E2]")
va2
```


Finally, when the parameter is the name of a species followed by
a `.` and the name of an event among `deaths`, `duplications`,
or `switches`,
<code>[TissueSimulation$var()](../reference/TissueSimulation-cash-var.md)</code>
returns the variable associated with the number of
the corresponding event in the specified species.

```{r}
# get the variable representing the number of epigenetic
# switches from A[E1]
va1_s <- sim$var("A[E1].switches")
va1_s

# get the variable representing the number of duplications in A+
sim$var("A[E1].duplications")

# get the variable representing the number of deaths in A+
sim$var("A[E1].deaths")
```

### Expressions and Formulas

An expression is one of the following objects:

-  a variable, e.g., `sim$var("A[E1]")`;
-  a numeric value, e.g., `3.4`;
-  the sum of two expressions, e.g., `sim$var("A[E1]") + 3.4`;
-  the subtraction of two expressions, e.g., `sim$var("A[E1]") - 3.4`;
-  the multiplication of two expressions, e.g., `sim$var("A[E1]") * 3.4`.

Two expression can be related by `<=`, `<`, `==`, `!=`, `>` and
`>=`.

A formula is:

-  a relation among two expressions, e.g., `sim$var("A[E1]")>=2`;
-  the conjunction of two formulas, e.g., `sim$var("A[E1]")>=2 & sim$var("A[E1]")<=500`;
-  the disjunction of two formulas, e.g., `sim$var("A[E1]")>=2 | sim$var("A[E1]")<=500`.

### The method <code>[TissueSimulation$run_until()](../reference/TissueSimulation-cash-run_until.md)</code>

The method <code>[TissueSimulation$run_until()](../reference/TissueSimulation-cash-run_until.md)</code>
takes a formula as a parameter and
runs the simulation until the formula no longer holds.

We can build a condition stating that the cardinality of `A[E1]` doubles that
of `A[E2]`
```{r}
c1 <- va1 >= 2 * va2
c1
```
..., a condition that holds when there are more than
100000 live cells of mutant `A`
```{r}
c2 <- va1 + va2 > 1e5
c2
```
..., a condition that constrains the simulated time
```{r}
# build a condition that holds when 40 time unit have been
# simulated at least
c4 <- v_time >= 40
c4
```
..., or deal with the number of species events. For instance, we can build a
condition that holds when less than 4000 epigenetic switches from the species
`A[E1]` have occurred.
```{r}
# build a condition that holds when less than 4000
# epigenetic switches from the species A[E1] have occurred
c3 <- va1_s < 4e3
c3
```

Moreover, we can logically combine different conditions.
```{r}
# build a condition that holds when c4 and at least one
# among c1, c2, and c3 hold
c5 <- c4 & (c1 | c2 | c3)
c5
```

Any simulation can be run until a condition holds.
```{r}
# place the initial cell
sim$place_cell("A[E1]", 500, 500)

# run the simulation while c5 does not hold
sim$run_until(c5)

sim
sim$get_clock()
```
