---
title: "Partitioning samples (FACS)"
output: rmarkdown::html_vignette
description: >
  How to partitioning samples, sorting cells, and simulate FACS
vignette: >
  %\VignetteIndexEntry{Partitioning samples (FACS)}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options:
  markdown:
    wrap: 72
bibliography: references.bib
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

> *Note:* This article presents advanced topics on sequencing simulation.
> Refer to `vignette("sequencing")` for an introduction on the subject.

> *Disclaimer:* RACES/ProCESS internally implements the
> probability distributions using the C++11 random number
> distribution classes. The standard does not specify their
> algorithms, and the class implementations are left free
> for the compiler. Thus, the simulation output depends on
> the compiler used to compile RACES, and because of that,
> the results reported in this article may differ from
> those obtained by the reader.


ProCESS allows to partition samples according to cell features as 
[FACS](https://www.sinobiological.com/category/fcm-facs-facs) does.
The method `PhylogeneticForest$partition_samples()` takes 
a labelling function, which accepts in input an object of the 
class `SampledCell` and returns a string, and applies it 
to the sampled cells. Then, it partitions the 
forest samples by using the obtained labels.
The resulting sub-samples are named according the format
`<sample name>_<cell label>`.

The method `PhylogeneticForest$partition_samples()` permanently changes the
phylogenetic forest from which is called. If users want to preserve the 
original forest, they must take care of it, for instance by copying or
saving it in a file.

```{r, echo=FALSE, results='hide', message=FALSE}
library(ProCESS)

phylo_forest <- load_phylogenetic_forest("phylo_forest.sff")
```

Let us assume to have built the phylogenetic forest `phylo_forest`
as detailed in `vignette("mutations")`. Information about the samples
from which the forest has been built can be obtained by using the
`PhylogeneticForest$get_samples_info()`.

```{r}
phylo_forest$get_samples_info()
```

The forest samples can be split by epigenetic state as it
follows.

```{r}
# the labelling function parameter has type `SampledCell`
epi_labelling <- function(cell) {

  # epigenetic state is "+", "-", "". However, "+" and "-"
  # are not supported in R data frame column names, so, we
  # can replace them by "P" and "N", respectively
  if (cell$epistate == "+") {
    return("P")
  }

  if (cell$epistate == "-") {
    return("N")
  }

  return("")
}

phylo_forest$partition_samples(epi_labelling)

phylo_forest$get_samples_info()
```

The same method can be easily applied to partition sampled cells
by mutant name...

```{r}
# loading again the original forest
phylo_forest <- load_phylogenetic_forest("phylo_forest.sff")

# a mutant-name-based labelling function
mutant_labelling <- function(cell) {
  return(cell$mutant)
}

phylo_forest$partition_samples(mutant_labelling)

phylo_forest$get_samples_info()
```

..., birth time...

```{r}
# loading again the original forest
phylo_forest <- load_phylogenetic_forest("phylo_forest.sff")

# a birth-time-based labelling function
birth_time_labelling <- function(cell) {
  if (cell$birth_time > 421) {
    return("YOUNG")
  }

  if (cell$birth_time > 321) {
    return("MIDDLE_AGED")
  }

  return("OLD")
}

phylo_forest$partition_samples(mutant_labelling)

phylo_forest$get_samples_info()
```

..., mutations...

```{r}
# loading again the original forest
phylo_forest <- load_phylogenetic_forest("phylo_forest.sff")

library(dplyr)

# collect all the sample passenger indels
passenger_indels <- phylo_forest$get_sampled_cell_mutations() %>%
  filter(class == "passenger", type == "indel")

# get one of the passenger indels
p_indel <- passenger_indels[sample(seq_len(nrow(passenger_indels)), 1), ]

p_indel

# a mutation-based labelling function that discriminates sampled cells
# containing indel `p_indel` from the other sampled cells
mutations_labelling <- function(cell) {
  has_indel <- nrow(cell$mutations %>%
                      filter(chr == p_indel[["chr"]],
                             from == p_indel[["from"]],
                             ref == p_indel[["ref"]],
                             alt == p_indel[["alt"]])) > 0

  if (has_indel) {
    return("HAS_MUTATION");
  }

  return("MISSES_MUTATION")
}

phylo_forest$partition_samples(mutant_labelling)

phylo_forest$get_samples_info()
```

... or combination of these properties.
