---
title: "Tissue sampling"
output: rmarkdown::html_vignette
description: >
  How to sample cells from a simulated tissue
vignette: >
  %\VignetteIndexEntry{Tissue sampling}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options:
  markdown:
    wrap: 72
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

> *Disclaimer:* ProCESS/CLONES implements probability distributions using the
> C++11 random number distribution classes. Since the standard does not specify
> the underlying algorithms, their implementations are left to the compiler.
> Consequently, the simulation output depends on the compiler used to build
> [CLONES](https://github.com/albertocasagrande/CLONES), and the results
> reported in this article may differ from those obtained by the reader.

Once one has familiarised oneself with how a tumour evolution simulation can be
programmed using ProCESS (see the [article on tissue simulation](tissue_simulation.md)), the next
step is to augment the simulation with sampling of tumour cells. This
mimics a realistic experimental design in which we collect tumour sequencing
data.

This article introduces sampling using different types of models; starting
from simpler to more complex simulation scenarios. We consider:

-   *multi-region sampling*: where at every time point multiple
    spatially-separated samples are collected;

-   *longitudinal sampling*: where the sampling is repeated at multiple
    time-points.

## Custom multi-region sampling

We consider a simple monoclonal model, without an epi-state.

```{r, eval=TRUE}
library("ProCESS")

# set the seed of the random number generator
set.seed(0)

# monoclonal model, no epigenetic state
sim <- TissueSimulation("Monoclonal")

sim$add_mutant("A", c(duplication = 0.1, death = 0.01))

sim$place_cell("A", 500, 500)
sim$run_up_to_size("A", 60000)
```

```{r,  fig.height=4, fig.width=3, fig.alt = c("A monoclonal tumour consisting of 60000 cells.")}
current <- plot_tissue(sim)
current
```

A sample is defined by a name and a bounding box, which has $(x,y)$
coordinates for the bottom-left and top-right points.

For this simulation, we define two samples with names `S_1_1` and
`S_1_2`.

```{r,  fig.height=4, fig.width=3, fig.alt=c("The rectangular regions corresponding to the samples S_1_1 and S_1_2.")}
# collect a squared box of (bbox_width x bbox_width) cells
bbox_width <- 50

# box A1
bbox1_p <- c(400, 400)
bbox1_q <- bbox1_p + bbox_width

# box B1
bbox2_p <- c(500, 500)
bbox2_q <- bbox2_p + bbox_width

library(ggplot2)

# view the boxes
current +
  geom_rect(xmin = bbox1_p[1], xmax = bbox1_q[2],
            ymin = bbox1_p[1], ymax = bbox1_q[2],
            fill = NA, color = "black") +
  geom_rect(xmin = bbox2_p[1], xmax = bbox2_q[2],
            ymin = bbox2_p[1], ymax = bbox2_q[2],
            fill = NA, color = "black")

# collect two samples from the two boxes
sim$sample_cells("S_1_1", bottom_left = bbox1_p, top_right = bbox1_q)
sim$sample_cells("S_1_2", bottom_left = bbox2_p, top_right = bbox2_q)
```

> *Note:* Sampling removes cells from the tissue, as if the tissue were
> surgically resected. Therefore, cells that are mapped to the bounding
> box after application of <code>[TissueSimulation$sample_cells()](../reference/TissueSimulation-cash-sample_cells.md)</code>
> are no longer part of the simulation.

A new call to `plot_tissue()` will show the box where the cells have
been removed to be white.

```{r,  fig.height=4, fig.width=3, fig.alt=c("The sampled tissue.")}
plot_tissue(sim)
```

This is also reflected by <code>[TissueSimulation$get_cells()](../reference/TissueSimulation-cash-get_cells.md)</code>,
which now will not find any tumour cells in the sampled region.

```{r}
library(dplyr)

# this should be empty
sim$get_cells(c(400, 400), c(400 + bbox_width, 400 + bbox_width)) %>% head
```

The sampling process exclusively collects tumour cells, while excluding
wild-type cells.

### The sample forest

Every sampled cell is linked, at the evolutionary level, to the other
cells that originate from the same initial cell. It helps to visualise
the evolutionary information on the cells that we have sampled as a
forest of trees (if one seeded multiple initial cells). The forest is an
object of the R6 class `SampleForest`.

```{r}
forest <- sim$get_sample_forest()

forest
```

The forest has methods to obtain the nodes of the sampled cells.

```{r}
forest$get_nodes() %>% head
```

The leaves of the forest are sampled cells, while the internal nodes are
their ancestors. The field `sample` is not available for internal nodes,
and reports the sample name otherwise.

```{r}
# the leaves in the forest represent sampled cells
forest$get_nodes() %>%
  filter(!is.na(.data$sample)) %>%
  head
```

The roots of the forest have no ancestors.

```{r}
# if it is one cell, then the forest is a tree
forest$get_nodes() %>%
  filter(is.na(.data$ancestor))
```

We can also query the forest about the samples used to build it.

```{r}
forest$get_samples_info()
```

We can visualise the forest. This plot reports the cells and, on the
y-axis, their time of birth.

```{r,  fig.height=6, fig.width=11, fig.alt=c("The sample forest of S_1_1 and S_1_2.")}
plot_forest(forest)
```

The plot also shows sample annotations and species, but for a large
number of cells, it can be difficult to view the full forest, unless a
very large canvas is used. For this reason, it is possible to subset
the forest.

```{r,  fig.height=6, fig.width=11, fig.alt=c("The sub-forest of sample S_1_1.")}
# extract the subforest linked to sample
S_1_1_forest <- forest$get_subforest_for("S_1_1")

plot_forest(S_1_1_forest)
```

In general, these plots can be annotated with extra information, such as
the sampling times and the MRCAs of each sample in the forest.

```{r,  fig.height=6, fig.width=11, fig.alt=c("The annotated forest of S_1_1 and S_1_2.","The annotated sub-forest of the sample S_1_1.")}
# full plot
plot_forest(forest) %>%
  annotate_forest(forest)
# S_1_1 plot
plot_forest(S_1_1_forest) %>%
  annotate_forest(S_1_1_forest)
```

## Randomised multi-region samples

```{r}
# set the seed of the random number generator
set.seed(0)

sim <- TissueSimulation("Randomised")

sim$add_mutant("A", c(duplication = 0.1, death = 0.01))

sim$place_cell("A", 500, 500)
sim$run_up_to_size("A", 60000)
```

We include a new mutant and let it grow. This new mutant has much higher
growth rates than its ancestor.

```{r,  fig.height=4, fig.width=3, fig.alt=c("A tumour having two mutants/species with different rates.")}
# add a new mutant
sim$add_mutant("B", c(duplication = 1, death = 0.01))
sim$mutate_progeny(sim$choose_border_cell_in("A"), "B")

sim$run_up_to_size("B", 10000)

current <- plot_tissue(sim)
current
```

ProCESS provides a
<code>[TissueSimulation$search_sample()](../reference/TissueSimulation-cash-search_sample.md)</code>
function to sample bounding boxes that contain a desired number of cells. The
function takes in input:

-   a bounding box size;
-   the number $n$ of cells to sample for a species of interest.

<code>[TissueSimulation$search_sample()](../reference/TissueSimulation-cash-search_sample.md)</code>
will attempt a fixed number of times to
sample the box, starting from positions occupied by the species of
interest. If a box that contains at least $n$ cells is not found within
a number of attempts, then the one with the largest number of samples is
returned.

This allows for programming sampling without having a clear idea of the
tissue conformation.

```{r,  fig.height=4, fig.width=3, fig.alt=c("The rectangle delimiting the sample S_2_1 containing 2000 cells of mutant B at least.")}
# a bounding box 50x50 with at least 100 cells of species B
n_w <- n_h <- 50
ncells <- 0.8 * n_w * n_h

# sampling ncells with random box sampling of boxes of size n_w x n_h
bbox <- sim$search_sample(c("B" = ncells), n_w, n_h)

# plot the bounding box
current +
  geom_rect(xmin = bbox$lower_corner[1], xmax = bbox$upper_corner[1],
            ymin = bbox$lower_corner[2], ymax = bbox$upper_corner[2],
            fill = NA, color = "black")
```

```{r}
# sample the tissue
sim$sample_cells("S_2_1", bbox$lower_corner, bbox$upper_corner)
```

Something similar to species `A`.

```{r,  fig.height=4, fig.width=3, fig.alt=c("The rectangle delimiting the sample S_2_2 containing 2000 cells of mutant A at least.")}
bbox <- sim$search_sample(c("A" = ncells), n_w, n_h)

# plot the bounding box
current +
  geom_rect(xmin = bbox$lower_corner[1], xmax = bbox$upper_corner[1],
            ymin = bbox$lower_corner[2], ymax = bbox$upper_corner[2],
            fill = NA, color = "black")
```

```{r}
# sample the tissue
sim$sample_cells("S_2_2", bbox$lower_corner, bbox$upper_corner)
```

The two samples have been extracted.

```{r,  fig.height=4, fig.width=3, fig.alt=c("The tissue of the biclonal tumour after the two samples have been collected.")}
plot_tissue(sim)
```

We can build the sample forest.

```{r,  fig.height=6, fig.width=11, fig.alt=c("The tissue of the biclonal tumour after the collection of the two samples.")}
forest <- sim$get_sample_forest()

plot_forest(forest) %>%
  annotate_forest(forest)
```

## Randomised cell sampling (Liquid biopsy)

ProCESS supports randomised cell sampling over the full tissue or a
rectangle thereof.

```{r, fig.alt=c("Two further samples S_2_3 and S_2_4 are collected. Their cells are randomly picked-up (like in liquid biopsy) from the whole tissue and from a bounding box, respectively. The figure depicts the sample forest of all the samples.")}
# collect up to 2500 tumour cells randomly selected over the whole tissue
sim$sample_cells("S_2_3", num_of_cells = 2500)

bbox <- sim$search_sample(c("A" = ncells), n_w, n_h)

# collect up to 200 tumour cells randomly selected in the provided
# bounding box
sim$sample_cells("S_2_4", bbox$lower_corner, bbox$upper_corner, 200)

forest <- sim$get_sample_forest()

plot_forest(forest) %>%
  annotate_forest(forest)
```

## Two populations with epigenetic states

We are now ready to simulate a model with epigenetic states and sub-clonal
expansions. The following example uses two epigenetic states. However, user can
create as many epigenetic states as they need.

```{r,  fig.height=4, fig.alt=c("A tumour having one mutants and two species with different rates.")}
# set the seed of the random number generator
set.seed(0)

sim <- TissueSimulation("Two Populations", epigenetic_states = c("E1", "E2"))

sim$death_activation_level <- 20

# add mutant "A" and set its species rates
sim$add_mutant("A", list(E1 = list(duplication = 0.1, death = 0.1, E2 = 0.01),
                         E2 = list(duplication = 0.08, death = 0.01,
                                   E1 = 0.01)))

sim$place_cell("A[E1]", 500, 500)
sim$run_up_to_size("A[E1]", 1000)
plot_tissue(sim, num_of_bins = 500)
```

We sample before introducing a new mutant.

```{r,  fig.height=4, fig.width=3, fig.alt=c("The sampled tumour.")}
bbox_width <- 10

sim$sample_cells("S_1_1",
                 bottom_left = c(480, 480),
                 top_right = c(480 + bbox_width, 480 + bbox_width))

sim$sample_cells("S_1_2",
                 bottom_left = c(500, 500),
                 top_right = c(500 + bbox_width, 500 + bbox_width))

plot_tissue(sim, num_of_bins = 500)
```

```{r,  fig.height=4, fig.width=3, fig.alt=c("The same tumour after 15 simulated time units have been elapsed.")}
# let the simulation evolve for further 15 time units
sim$run_up_to_time(sim$get_clock() + 15)

plot_tissue(sim, num_of_bins = 500)
```

Add a new submutant.

```{r,  fig.height=4, fig.width=3, fig.alt=c("A new mutant with epigenetic state arose.")}
cell <- sim$choose_border_cell_in("A")

sim$add_mutant("B", list(E1 = list(duplication = 0.08, death = 0.05, E2 = 0.05),
                         E2 = list(duplication = 0.3, death = 0.05, E1 = 0.1)))

sim$mutate_progeny(cell, "B")

# let it grow more
sim$run_up_to_size("B[E1]", 7000)

plot_tissue(sim, num_of_bins = 500)
```

Sample again and plot the tissue

```{r,  fig.height=4, fig.width=3, fig.alt=c("The tumour tissue after two samples have been collected.","The Muller plot of the biclonal tumour with epigenetic states.")}
n_w <- n_h <- 25
ncells <- 0.9 * n_w * n_h

bbox <- sim$search_sample(c("A" = ncells), n_w, n_h)
sim$sample_cells("S_2_1", bbox$lower_corner, bbox$upper_corner)

bbox <- sim$search_sample(c("B" = ncells), n_w, n_h)
sim$sample_cells("S_2_2", bbox$lower_corner, bbox$upper_corner)

plot_tissue(sim, num_of_bins = 500)
plot_muller(sim)
```

Now we show the sample forest, which starts being rather
complicated

```{r,  fig.height=12, fig.width=17, fig.alt=c("The sample forest of the biclonal tumour with epigenetic states.")}
forest <- sim$get_sample_forest()

plot_forest(forest) %>%
  annotate_forest(forest)
```

### Storing Samples Forests

A sample forest can be saved in a file by using the method
<code>[SampleForest$save()](../reference/SampleForest-cash-save.md)</code>.

```{r, echo=FALSE,  results='hide', message=FALSE}
unlink("sample_forest.sff")
```

```{r}
# check the file existence. It should not exist.
file.exists("sample_forest.sff")

# save the sample forest in the file "sample_forest.sff"
forest$save("sample_forest.sff")

# check the file existence. It now exists.
file.exists("sample_forest.sff")
```

The saved sample forest can successively be loaded by using the function
`load_forest()`.

```{r}
# load the sample forest from "sample_forest.sff" and store it in `forest2`
forest2 <- load_forest("sample_forest.sff")

# let us now compare the sample forests stored in `forest` and `forest2`;
# they should be the same.
forest
forest2
```
