---
title: "Partitioning samples (FACS)"
output: rmarkdown::html_vignette
description: >
  How to partitioning samples, sorting cells, and simulate FACS
vignette: >
  %\VignetteIndexEntry{Partitioning samples (FACS)}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options:
  markdown:
    wrap: 72
bibliography: references.bib
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

> *Note:* This article presents advanced topics on sequencing simulation.
> Refer to [this article](sequencing.md) for an introduction on the subject.

> *Disclaimer:* ProCESS/CLONES implements probability distributions using the
> C++11 random number distribution classes. Since the standard does not specify
> the underlying algorithms, their implementations are left to the compiler.
> Consequently, the simulation output depends on the compiler used to build
> [CLONES](https://github.com/albertocasagrande/CLONES), and the results
> reported in this article may differ from those obtained by the reader.


ProCESS allows to partition samples according to cell features as
[FACS](https://www.sinobiological.com/category/fcm-facs-facs) does.
The method <code>[PhylogeneticForest$partition_samples()](../reference/PhylogeneticForest-cash-partition_samples.md)</code>
takes a labelling function, which accepts as input an object of the class
<code>[PhylogeneticForestNode](../reference/PhylogeneticForestNode.md)</code>
and returns a string, and applies it to the sampled cells. Then,
it partitions the forest samples by using the obtained labels. The resulting
sub-samples are named according the format `<sample name>_<cell label>`.

The method <code>[PhylogeneticForest$partition_samples()](../reference/PhylogeneticForest-cash-partition_samples.md)</code>
permanently changes the phylogenetic forest from which is called. If users want
to preserve the original forest, they must take care of it, for instance by
copying or saving it in a file.

```{r, echo=FALSE, results='hide', message=FALSE}
library("ProCESS")

phylo_forest <- load_forest("phylo_forest.pff", quiet = TRUE)
```

Let us assume to have built the phylogenetic forest `phylo_forest` as detailed
in [this article](mutations.md). Information about the samples from which the
forest has been built can be obtained by using the
<code>[PhylogeneticForest$get_samples_info()](../reference/PhylogeneticForest-cash-get_samples_info.md)</code>.

```{r}
phylo_forest$get_samples_info()
```

The forest samples can be split by epigenetic state as it
follows.

```{r}
# the labelling function parameter has type `PhylogeneticForestNode`
epi_labelling <- function(forest_node) {
  forest_node$epistate_name
}

phylo_forest$partition_samples(epi_labelling)

phylo_forest$get_samples_info()
```

The same method can be easily applied to partition sampled cells
by mutant name...

```{r}
# loading again the original forest
phylo_forest <- load_forest("phylo_forest.pff")

# a mutant-name-based labelling function
mutant_labelling <- function(forest_node) {
  forest_node$mutant_name
}

phylo_forest$partition_samples(mutant_labelling)

phylo_forest$get_samples_info()
```

..., birth time...

```{r}
# loading again the original forest
phylo_forest <- load_forest("phylo_forest.pff")

# a birth-time-based labelling function
birth_time_labelling <- function(forest_node) {
  if (forest_node$birth_time > 421) {
    return("YOUNG")
  }

  if (forest_node$birth_time > 321) {
    return("MIDDLE_AGED")
  }

  return("OLD")
}

phylo_forest$partition_samples(birth_time_labelling)

phylo_forest$get_samples_info()
```

..., mutations...

```{r}
# loading again the original forest
phylo_forest <- load_forest("phylo_forest.pff")

library(dplyr)

# collect all the sample passenger indels
passenger_indels <- phylo_forest$get_sampled_cell_mutations() %>%
  filter(nature == "passenger", nchar(ref) != nchar(alt))

# get one of the passenger indels and build the mutation
p_indel <- passenger_indels[sample(seq_along(passenger_indels), 1), ]

mutation <- Mutation(chr = p_indel[["chr"]], from = p_indel[["from"]],
                     ref = p_indel[["ref"]], alt = p_indel[["alt"]])

mutation

# find the nodes where the mutation occurs for the first time
first_node_ids <- phylo_forest$get_first_occurrences(mutation)

# a mutation-based labelling function that discriminates sampled cells
# that are descendant of the first nodes in which the mutation occurs, i.e.,
# sampled cells containing indel `p_indel`, from the other sampled cells
mutations_labelling <- function(forest_node) {

  while (forest_node$cell_id < first_node_ids) {
    forest_node <- forest_node$parent
  }

  if (forest_node$cell_id == first_node_ids) {
    return("HAS_MUTATION");
  }

  return("MISSES_MUTATION")
}

phylo_forest$partition_samples(mutations_labelling)

phylo_forest$get_samples_info()
```

... or combination of these properties.
