---
title: "Time Series Analysis"
output: rmarkdown::html_vignette
description: >
  Analyse the evolution of the system over time
vignette: >
  %\VignetteIndexEntry{Time Series Analysis}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options:
  markdown:
    wrap: 72
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```


```{r echo=FALSE, results='hide', message=FALSE, warning=FALSE}
library(ProCESS)

# set the seed of the random number generator
set.seed(0)

# create a tissue simulation with epigenetic states whose tissue has
# size 1200x900
sim <- TissueSimulation("Test", epigenetic_states = c("E1", "E2"),
                        width = 1200, height = 900)

sim$death_activation_level <- 50

# add a mutant "A"
sim$add_mutant("A", list(E1 = list(duplication = 0.2, death = 0.1, E2 = 0.01),
                         E2 = list(duplication = 0.08, death = 0.01,
                                   E1 = 0.01)))

# we add one cell of species A[E1] (mutant A in epi-state E1) in
# position (500, 500).
sim$place_cell("A[E1]", 500, 500)

# run the simulation until the species A[E1] consists of less than 1000 cells
sim$run_up_to_size("A[E1]", 1000)

# run the simulation until less than 100 deaths has occurred in species A[E2]
sim$run_up_to_event("death", "A[E2]", 100)

# run the simulation for other 15 time units
sim$run_up_to_time(sim$get_clock() + 15)

# add a mutant "B"
sim$add_mutant("B", list(E1 = list(duplication = 0.3, death = 0.05, E2 = 0.05),
                         E2 = list(duplication = 0.3, death = 0.1,
                                   E1 = 0.05)))

# mutant injection
sim$mutate_progeny(sim$choose_border_cell_in("A"), "B")

# run the simulation until there are less than cells in B[E1]
sim$run_up_to_size("B[E1]", 600)

# define evolutionary parameters
sim$add_mutant("C", list(E1 = list(duplication = 0.2, death = 0.1, E2 = 0.1),
                         E2 = list(duplication = 0.4, death = 0.01,
                                   E1 = 0.1)))

# choose and mutate
sim$mutate_progeny(sim$choose_border_cell_in("A"), "C")

sim$run_up_to_time(sim$get_clock() + 7)
```

> *Note:* This article presents advanced topics on tissue simulation. Refer to
> [this article](tissue_simulation.md) for an introduction on the subject.

> *Disclaimer:* ProCESS/CLONES implements probability distributions using the
> C++11 random number distribution classes. Since the standard does not specify
> the underlying algorithms, their implementations are left to the compiler.
> Consequently, the simulation output depends on the compiler used to build
> [CLONES](https://github.com/albertocasagrande/CLONES), and the results
> reported in this article may differ from those obtained by the reader.

Sometimes it is convenient to plot a time series of a simulation,
reporting species or firing counts over time. Since ProCESS is
programmable, it is easy to make a for-loop algorithm and collect
the simulation data over time.

## Default History-Based Data

However, this is not required because at the end of any `run_to_*`
method, CLONES stores data on the number of species cells and
the number of event firings. These data can be extracted.

Let us consider the simulation `sim` as produced in
[this article](tissue_simulation.md).

```{r}
# to use %>% the package dplyr is required
library(dplyr)

# the firings
sim$get_firing_history() %>% head()

# for example, total number of the deaths on `B+` at the end of the
# previous calls of the `run_to_*` methods
sim$get_firing_history() %>%
  filter(event == "death", mutant == "B", epistate == "E2")

# the counts
sim$get_count_history() %>% head()
```

The time-series can be plot using `plot_timeseries()`

```{r,  fig.height=4, fig.width=4, fig.alt=c("A time-series plot. It presents the cell count per species along the simulation.")}
# time-series plot
plot_timeseries(sim)
```

## Custom Time-Series

If the default time-series is not enough coarse-grained, one can set
<code>[TissueSimulation$history_delta](../reference/TissueSimulation-cash-history_delta.md)</code>
to increase the sampling rate of
the state (by default, <code>[TissueSimulation$history_delta](../reference/TissueSimulation-cash-history_delta.md)</code> is set to $0$).

We show this by re-simulating a tumour with two sub-mutants.

```{r}
# example time-series on a new simulation, with coarse-grained time-series
sim <- TissueSimulation("Finer Time Series", epigenetic_states = c("E1", "E2"))

sim$history_delta <- 1
sim$death_activation_level <- 100

# add a mutant "A"
sim$add_mutant("A", list(E1 = list(duplication = 0.2, death = 0.1, E2 = 0.01),
                         E2 = list(duplication = 0.08, death = 0.01,
                                   E1 = 0.01)))

sim$place_cell("A[E1]", 500, 500)
sim$run_up_to_size("A[E2]", 400)

# add a mutant "B"
sim$add_mutant("B", list(E1 = list(duplication = 0.3, death = 0.3, E2 = 0.05),
                         E2 = list(duplication = 0.3, death = 0.1, E1 = 0.05)))

sim$mutate_progeny(sim$choose_border_cell_in("A"), "B")

sim$run_up_to_size("B[E2]", 300)

# add a mutant "C"
sim$add_mutant("C", list(E1 = list(duplication = 0.2, death = 0.1, E2 = 0.1),
                         E2 = list(duplication = 0.4, death = 0.01, E1 = 0.1)))

sim$mutate_progeny(sim$choose_border_cell_in("B"), "C")

# add a mutant "D" - has the same rates of "C"
sim$add_mutant("D", list(E1 = list(duplication = 0.2, death = 0.1, E2 = 0.1),
                         E2 = list(duplication = 0.4, death = 0.01, E1 = 0.1)))

sim$mutate_progeny(sim$choose_border_cell_in("A"), "D")

sim$run_up_to_size("D[E1]", 100)
```

The time-series can be plot using `plot_timeseries()`.

```{r,  fig.height=4, fig.width=4, fig.alt=c("A more accurate time-series plot. The count sampling time is configurable.", "The same time-series plot can be presented by using a log-scale.")}
# time-series plot
plot_timeseries(sim)

# logscale helps seeing the different effective growth rates
plot_timeseries(sim) + ggplot2::scale_y_log10()
```

### Muller Plot

We can also get a Muller plot of the evolution using
[ggmuller](https://CRAN.R-project.org/package=ggmuller).

```{r,  fig.height=4, fig.width=6, message=FALSE, warning=FALSE, fig.alt=c("A Muller plot of the simulation. It depicts the ratios between species along the simulation.")}
# default Muller plot
plot_muller(sim)
```

In this case every population is annotated as a descendant of the
ancestor mutant. Note however that reversible epi-states do not fit a
traditional Muller plot because they violate the no-back mutation model.

In this case, _ProCESS_ will show first the epi-state that was randomly
injected in the simulation, and the second will result by linear. This
is not a completely correct perspective of the simulation time-series;
still, it help understand trends.

```{r,  fig.height=4, fig.width=6, message=FALSE, warning=FALSE, fig.alt=c("The Muller plot for the last quater of the simulation.", "The Muller plot in log-scale for the last quater of the simulation.")}
# custom Mullers
clock <- sim$get_clock()

plot_muller(sim) + ggplot2::xlim(clock * 3/4, clock)

plot_muller(sim) +
  ggplot2::xlim(clock * 3/4, clock) +
  ggplot2::scale_y_log10()
```

## Time-Varying Evolutionary Rates

The rates of any species may change over time. For instance, this
happens when a population is subject to a targeted treatment.

For instance, going back to the example simulation, we can increase the death
rates of `C[E1]`, `C[E2]`, `B[E1]`, and `B[E2]`.

```{r,  fig.height=4, fig.width=6, message=FALSE, warning=FALSE, fig.alt=c("The Muller plot for the last quarter of the simulation after changing the death rates of B and C species.", "The Muller plot in log-scale  for the last quater of the simulation after changing the death rates of B and C species.")}
# current rates
sim

# raise the death rate levels of "B" species
sim$set_rate("B[E1]", "death", 0.5)
sim$set_rate("B[E2]", "death", 0.5)

# ... and those of "C" species
sim$set_rates(list(C = list(E1 = list(death = 0.5),
                            E2 = list(death = 0.5))))

# now D will grow
sim$run_up_to_size("D[E1]", 6000)

# current state
sim

# this now show the change in rates
clock <- sim$get_clock()
plot_muller(sim) + ggplot2::xlim(clock * 3 / 4, clock)

plot_muller(sim) +
  ggplot2::xlim(clock * 3/4, clock) +
  ggplot2::scale_y_log10()
```

