---
title: "Tissue simulation"
output: rmarkdown::html_vignette
description: >
  Learn how to simulate tumour evolution
vignette: >
  %\VignetteIndexEntry{Tissue simulation}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options:
  markdown:
    wrap: 72
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

> *Disclaimer:* ProCESS/CLONES internally implements the
> probability distributions using the C++11 random number
> distribution classes. The standard does not specify their
> algorithms, and the class implementations are left free
> for the compiler. Thus, the simulation output depends on
> the compiler used to compile [CLONES](https://github.com/albertocasagrande/CLONES), and
> because of that, the results reported in this article may
> differ from those obtained by the reader.

The following steps are required to simulate a tissue.

1.  creation of a tissue;

2.  introduction of cells in the tissue;

3.  actual simulation.

The simulation is managed by an object of the R6 class `TissueSimulation`,
which allows programmingm the tissue evolution over time, adding new cells
as the simulation progresses. The state of the simulation and
tissue can be visualised using `ggplot2`-powered plots.

## Tissue specification

To perform a simulation, a new object of class `TissueSimulation` must be
created.

```{r, eval=FALSE}
library(ProCESS)

# Default constructor
sim <- TissueSimulation()
```

This automatically builds a 1000x1000-cell tissue and sets the name of the 
simulation to be `clones_<date>_<hour>`.

A simulation custom name can be specified as it follows.

```{r, eval = TRUE}
library(ProCESS)

# set the seed of the random number generator
set.seed(0)

# Call your simulation "Test"
sim <- TissueSimulation("Test")
```

The `set.seed()` call is not mandatory and can be omitted; however, it is
highly recommended to ensure the simulation repeatability.

The optional Boolean parameter `save_snapshots` can be used to save the 
simulation progress to the disk and recover it later. By
setting `save_snapshots` to `TRUE`, the simulation progress os
saved in a directory named after the simulation.

```{r, eval = FALSE}
# The progress of the simulation will be saved in the "Test" folder.
# If the directory "Test" already exists, an exception is raised.
sim <- TissueSimulation("Test", save_snapshots = TRUE)
```

```{r echo=FALSE, results='hide', message=FALSE, warning=FALSE}
sim$death_activation_level <- 50
```

We can get information about any `TissueSimulation` object by printing the
corresponding variable.

```{r, eval}
sim
```

`TissueSimulation` objects expose methods to retrieve specific information 
and control the modelled simulations.

```{r}
# Get the simulation directory, i.e., "Test"
sim$get_name()

# Get the tissue size, i.e., c(1000,1000)
sim$get_tissue_size()
```

### Custom Tissue Sizes

The tissue size can be specified during simulation creation using
the parameters `width` and `height`.

```{r}
# build a tissue simulation whose tissue has a width of 1200 and
# a height of 900
sim <- TissueSimulation("Test", width = 1200, height = 900)

# Get the tissue size, i.e., c(1200,900)
sim$get_tissue_size()
```

## Species specification

In order to simulate the evolution of some species, we need to add them
to `sim`. This process defines the evolutionary parameters of the
species.

A *mutant* is a set of cells having the same (potentially unknown) driver
mutations. Cells in the same mutant can have different liveness rates
due to different epigenetic states.

A *species* is a mutant with an optional epigenetic state. At this point in the
simulation, the mutant is just a name (`A`, `B`, etc.) that could later
be linked to mutations of interest. The epigenome is a binary
feature of a species, represented by epi-states `+`/`-` (positive
and negative status). This is an abstraction that could represent an
active/inactive state linked to promoter methylation or, more broadly,
a phenotype. The evolution of mutants is non-reversible (no-back
mutations model), while the evolution among epi-states is potentially
reversible.

For example, if we define two mutants `A` and `B` with their epi-states
`+`/`-`, we have 4 distinct species: `A+` and `A-`, as well as `B+` and
`B-`.

Hybrid models can be obtained, e.g., a mutant `A` (with no epi-states),
together with `B+` and `B-`.

### Evolutionary parameters

We use a notation common in linear birth-death processes. If a species
`A` has no epi-states, then its stochastic behaviour is defined by the
state-change rates

$$
\begin{align}
\text{(growth)}\quad A & \rightarrow_{\lambda} 2 A \\
\text{(death)}\quad A & \rightarrow_{\delta} \emptyset
\end{align}
$$

where:

-   $\lambda>0$ is the cell growth rate;
-   $\delta>0$ is the cell death rate.

Instead, if the species has epi-state `+` (denoted $A_\oplus$) and `-`
(denoted $A_\ominus$), then its stochastic behaviour is defined by the
state-change rates

$$
\begin{align}
\text{(growth +)}\quad A_\oplus & \rightarrow_{\lambda_\oplus} 2 A_\oplus \\
\text{(death +)}\quad A_\oplus & \rightarrow_{\delta_\oplus} \emptyset \\
\text{(growth -)}\quad A_\ominus & \rightarrow_{\lambda_\ominus} 2 A_\ominus \\
\text{(death -)}\quad A_\ominus & \rightarrow_{\delta_\ominus} \emptyset \\
\text{(switch +-)}\quad A_\oplus & \rightarrow_{\epsilon_{+-}} A_\oplus + A_\ominus \\
\text{(switch -+)}\quad A_\ominus & \rightarrow_{\epsilon_{-+}} A_\ominus + A_\oplus
\end{align}
$$

where the rates $\lambda$ and $\delta$ are as above, and $\epsilon_{+-}$
or $\epsilon_{-+}$ are the rates at which cells of a certain epi-state
duplicate and flip the epigenetic marker of one of the progeny.

```{r, eval=TRUE}
sim$add_mutant(name = "A",
               epigenetic_rates = c("+-" = 0.01, "-+" = 0.01),
               growth_rates = c("+" = 0.2, "-" = 0.08),
               death_rates = c("+" = 0.05, "-" = 0.01))

# updated object (counts refer to number of cells of each species)
sim

# Get the simulation species and their rates
sim$get_species()
```

A mutant without epi-states could be added as

```{r, eval=FALSE}
# Not run
sim$add_mutant(name = "A", growth_rates = 0.2, death_rates =  0.1)
```

To be able to simulate the model, an initial cell needs to be displaced
in the tissue.

```{r}
# We add one cell of species A+ (mutant A in epi-state +) in
# position (500, 500).
sim$place_cell("A+", 500, 500)
```

### Visualisations

We can query the current state of the simulation and extract the
position of each cell in the tissue.

```{r}
# Counts per species
sim$get_counts()

# Cells position (one so far)
sim$get_cells()
```

This information can be plotted. Note that the tissue visualisation uses
hexagonal bins to avoid plotting delays when the simulation uses
thousands of cells.

```{r,  fig.height=4, fig.width=3, fig.alt=c("A pie-chart of the species cell count of a tumour consisting of a single cell.", "A representation of the tissue consisting of a single cell.")}
# Pie-chart for counts
plot_state(sim)

# Spatial distribution for the whole tissue (hexagonal bins)
plot_tissue(sim)
```

> **Note:** since the plots are done with `ggplot2` they can be assembled
> and customised.

## Species Evolution

There are 4 ways to let the simulation evolve:

-   advancing until the number of cells in a species reaches a given
    threshold $\theta > 0$;
-   advancing until a new time $t>0$ is reached;
-   advancing until a desired number of firings (of one particular
    event) has occurred;
-   advancing until a formula is not satisfied by the simulation
    status (_advanced topic_; see [this article](run_until.html)).

### Size-Based Simulation

We can run the simulation up to when we have $\theta > 0$; cells of species `A+`

```{r,  fig.height=4, fig.width=3, fig.alt=c("The tissue of a tumour with one mutant `A` and two species `A+` and `A-`. The species `A+` consists of 500 cells.")}
# Counts per species is now 0
sim$get_counts()

sim$run_up_to_size("A+", 500)

# Counts per species now reports 500 for A+
sim$get_counts()

plot_tissue(sim)
```

### Firing-Based Simulation

The number of times each event has fired is accessible

```{r}
# Get the number of fired event per species
sim$get_firings()
```

A small number of cell deaths have occurred in species `A-` up to this
point, so we can simulate the system until there are 100 of them.

```{r}
sim$run_up_to_event("death", "A-", 100)

# The row "death", for "A" "-" now reports 100
sim$get_firings()
```

```{r,  fig.height=4, fig.width=3, fig.alt=c("The tumour tissue after letting the simulation evolves until 100 deaths occurred to `A-` cells.")}
# Plot the tissue by using 200 bins
plot_tissue(sim, num_of_bins = 200)
```

### Clock-Based Simulation

It is also possible to take the current simulation clock as a reference,
and simulate further.

```{r}
# Get the simulation clock
sim$get_clock()

# Run the simulation for other 15 time units
sim$run_up_to_time(sim$get_clock() + 15)

# Get again the simulation clock
sim$get_clock()
```

```{r,  fig.height=4, fig.width=3, fig.alt=c("The tumour tissue after letting the simulation evolves until 100 deaths occurred to `A-` cells.")}
plot_tissue(sim)
```

##### Getting Cells (Advanced)

At this point, if we query the simulation, we will find more cells (we
use `dplyr` to process query results). For convenience, the getters
accept parameters to subset the tissue.

```{r, message=FALSE}
# load dplyr to use %>%
require(dplyr)

# Get the cells in the tissue at current simulation time
sim$get_cells() %>% head()

# Get the cells in the tissue rectangular sample having
# [500,500] and [505,505] as lower and upper corners, respectively
sim$get_cells(c(500, 500), c(505, 505))  %>% head()

# Get the cells in the tissue having epigenetic state "-"
sim$get_cells(c("A", "B"), c("-")) %>% head()

# Get the cells in the tissue having epigenetic state "-" and,
# at the same time, belonging to rectangular sample bounded by
# [500,500] and [505,505] as lower and upper corners, respectively
sim$get_cells(c(500, 500), c(505, 505), c("A", "B"), c("-"))  %>% head()
```

### Evolving new species

ProCESS can select cells from the tissue randomly for each mutant, or
within a constrained tissue area.

```{r}
# Stochastic sampling from the whole tissue: it can return A+ or A-
sim$choose_cell_in("A")

# Calling it again may result in a different cell
sim$choose_cell_in("A")

# Constrain sampling in the tissue rectangular selection [500,550]x[350,450]
sim$choose_cell_in("A",  c(500, 350), c(550, 450))
```

This feature can be used to program the generation of new species,
mimicking new mutants that generate subclonal expansions.

Imagine we want to add a new mutant `B` with epi-states -- and
therefore new species `B+` and `B-` -- as descending from `A`, we need to:

-   locate one cell of mutant `A` in the tissue, which is where we
    will inject the new mutant;

-   add the specifics of mutant `B` (via <code>[TissueSimulation$add_mutant()](../reference/TissueSimulation-cash-add_mutant.html)</code>, 
    as we did for `A`);

-   implement the change of the cell of mutant `A` to a cell of
    mutant `B`.

```{r}
# We locate a random cell
cell <- sim$choose_cell_in("A")
cell

# Add mutant
sim$add_mutant(name = "B",
               epigenetic_rates = c("+-" = 0.05, "-+" = 0.05),
               growth_rates = c("+" = 0.7, "-" = 0.3),
               death_rates = c("+" = 0.05, "-" = 0.1))
```

Then we inject the cell, and simulate a little bit.

```{r}
# Mutant injection
sim$mutate_progeny(cell, "B")

# Generated event A -> B either in + or - epi-states is now recorded
sim$get_counts()

# New evolution
sim$run_up_to_size("B+", 20000)
```

At this point, we can inspect the tissue in more detail. It can help to
face the species to clearly appreciate the spatial diffusion of the
populations.

```{r,  fig.height=4, fig.width=3, fig.alt=c("The species count after the new mutant `B` arose.", "The tumour tissue after the new mutant `B` arose.")}
plot_state(sim)
plot_tissue(sim, num_of_bins = 250)
```

```{r,  fig.height=6, fig.width=6, fig.alt=c("The tumour tissue in which cells are split by species.")}
# Facet on species via ggplot
library(ggplot2)

plot_tissue(sim, num_of_bins = 250) + facet_wrap(~species)
```

If, at this point in the simulation, we generate a new mutant `C` from
`A` in the rectangle $[450,500]\times [550, 600]$.

```{r}
# Define evolutionary parameters
sim$add_mutant(name = "C",
               epigenetic_rates = c("+-" = 0.1, "-+" = 0.1),
               growth_rates = c("+" = 0.2, "-" = 0.4),
               death_rates = c("+" = 0.1, "-" = 0.01))

# Choose and mutate
sim$mutate_progeny(sim$choose_cell_in("A",  c(450, 550), c(500, 600)), "C")

sim$run_up_to_time(sim$get_clock() + 7)
```

```{r,  fig.height=4, fig.width=3, fig.alt=c("The species count after the third mutant `C` arose.", "The tumour tissue after the third mutant `C` mutant arose.")}
plot_state(sim)
plot_tissue(sim, num_of_bins = 250)
```

### Other Operations

##### Injection of Cells into a Tissue

In the tissue, we can manually inject multiple cells; all injected cells
can be retrieved.

```{r}
# Now it will return just the initial cell
sim$get_added_cells()
```

##### Avoiding Drift

Any species with a non-zero death rate can become extinct
due to *drift*.

Drift makes it difficult to simulate and be confident about what species are
in the model. To facilitate the user,
[CLONES](https://github.com/albertocasagrande/CLONES) can avoid drift by
setting a *death activation level*. This value is the minimum number of cells
that enables cell death in a species: a cell of species $S$ can die if and
only if $S$ has reached the death activation level at least once during
the simulation.

This threshold applies to all species and is set to 1 by default.

```{r}
sim$death_activation_level

# Change death activation level
sim$death_activation_level <- 50
```
