---
title: "Tissue simulation"
output: rmarkdown::html_vignette
description: >
  Learn how to simulate tumour evolution
vignette: >
  %\VignetteIndexEntry{Tissue simulation}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options:
  markdown:
    wrap: 72
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

> *Disclaimer:* ProCESS/CLONES implements probability distributions using the
> C++11 random number distribution classes. Since the standard does not specify
> the underlying algorithms, their implementations are left to the compiler.
> Consequently, the simulation output depends on the compiler used to build
> [CLONES](https://github.com/albertocasagrande/CLONES), and the results
> reported in this article may differ from those obtained by the reader.

The following steps are required to simulate a tissue.

1.  creation of a tissue;

2.  introduction of cells in the tissue;

3.  actual simulation.

The simulation is managed by an object of the S4 class `TissueSimulation`,
which allows programming the tissue evolution over time, adding new cells
as the simulation progresses. The state of the simulation and
tissue can be visualised using `ggplot2`-powered plots.

## Tissue specification

To perform a simulation, a new object of class `TissueSimulation` must be
created.

```{r, eval=FALSE}
library(ProCESS)

# default constructor
sim <- TissueSimulation()
```

This automatically builds a 1000x1000-cells tissue - which can host one
million cells - and sets the name of the simulation to be
`clones_<date>_<hour>`.

A simulation custom name can be specified as it follows.

```{r, eval = TRUE}
library(ProCESS)

# set the seed of the random number generator
set.seed(0)

# call your simulation "Test"
sim <- TissueSimulation("Test")
```

The `set.seed()` call is not mandatory and can be omitted; however, it is
highly recommended to ensure the simulation repeatability.

The optional Boolean parameter `save_snapshots` can be used to save the
simulation progress to the disk and recover it later. By
setting `save_snapshots` to `TRUE`, the simulation progress os
saved in a directory named after the simulation.

```{r, eval = FALSE}
# the progress of the simulation will be saved in the "Test" folder.
# If the directory "Test" already exists, an exception is raised.
sim <- TissueSimulation("Test", save_snapshots = TRUE)
```

```{r echo=FALSE, results='hide', message=FALSE, warning=FALSE}
sim$death_activation_level <- 50
```

We can get information about any `TissueSimulation` object by printing the
corresponding variable.

```{r, eval}
sim
```

`TissueSimulation` objects expose methods to retrieve specific information
and control the modelled simulations.

```{r}
# get the simulation directory, i.e., "Test"
sim$get_name()

# get the tissue size, i.e., c(1000,1000)
sim$get_tissue_size()
```

### Custom Tissue Sizes

The tissue size can be specified during simulation creation using
the parameters `width` and `height`.

```{r}
# build a tissue simulation whose tissue has a width of 1200 and
# a height of 900
sim <- TissueSimulation("Test", width = 1200, height = 900)

# get the tissue size, i.e., c(1200,900)
sim$get_tissue_size()
```

## Species specification

In order to simulate the evolution of some species, we need to add them
to `sim`. This process defines the evolutionary parameters of the
species.

A *mutant* is a set of cells having the same (potentially unknown) driver
mutations. Cells in the same mutant can have different liveness rates
due to different epigenetic states.

A *species* is a mutant with an optional epigenetic state. At this point in the
simulation, the mutant is just a name (`A`, `B`, etc.) that could later be
linked to mutations of interest. The epigenomic state is a discrete feature of a
species. This is an abstraction that could represent a set of active/inactive
states linked to promoter methylations or, more broadly, a phenotype. The
evolution of mutants is non-reversible (no-back mutations model), while the
evolution among epigenetic states is potentially reversible.

For example, if we define a simulation with three epigenetic states `E1`, `E2`,
and `E3` and two mutants `A` and `B`, we have 6 distinct species: `A[E1]`,
`A[E2]`, `A[E3]`, `B[E1]`, `B[E2]`, and `B[E3]`.

The tissue epigenetic states can be added either at the simulation construction
or by using the method
<code>[TissueSimulation$add_epigenetic_states()](../reference/TissueSimulation-cash-add_epigenetic_states.md)</code>.

```{r}
sim <- TissueSimulation(epigenetic_states = c("E1", "E2"))

# add two further epigenetic states
sim$add_epigenetic_states(c("E3", "E4"))

# get the simulation epigenetic states
sim$get_epigenetic_states()
```

### Evolutionary parameters

We use a notation common in linear birth-death processes. If a simulation has
no epigenetic states, then any mutant $A$ has exactly one species $A$ and its
the stochastic behaviour is defined by the state-change rates

$$
\begin{aligned}
\text{(duplication)}\quad A & \rightarrow_{\lambda_{A}} 2 A \\
\text{(death)}\quad A & \rightarrow_{\delta_{A}} \emptyset
\end{aligned}
$$

where:

-   $\lambda_{A}>0$ is the cell duplication rate in $A$;
-   $\delta_{A}>0$ is the cell death rate in $A$.

Instead, if the simulation has $n$ epigenetic states $E_1$, $E_2$,..., $E_n$,
then the species of the mutant $A$ has $n$ species denoted $A[E_1]$, $A[E_2]$,
..., $A[E_n]$, respectively. The stochastic behaviour of the species $A[E_i]$ is
defined by the state-change rates

$$
\begin{aligned}
\text{(duplication $E_i$)}\quad A[E_i] & \rightarrow_{\lambda_{A,i}} 2 A[E_i] \\
\text{(death $E_i$)}\quad A[E_i] & \rightarrow_{\delta_{A,i}} \emptyset \\
\text{(switch from $E_i$ to $E_1$)}\quad A[E_i] & \rightarrow_{\epsilon_{A,i,1}} A[E_1] + A[E_i] \\
&\vdots\\
\text{(switch from $Ei$ to $E_n$)}\quad A[Ei] & \rightarrow_{\epsilon_{A,i,n}} A[E_n] + A[E_i]
\end{aligned}
$$

where the rates $\lambda_{A,i}$ and $\delta_{A,i}$ are as above, and
$\epsilon_{A,i,j}$ is the rate at which cells of the species $A[E_i]$ duplicate
and let one of its progeny switch to $A[E_j]$.

Let us add a mutant to the simulation with epigenetic states `E1`, `E2`, `E3`,
and `E4` by using the method
<code>[TissueSimulation$add_mutant()](../reference/TissueSimulation-cash-add_mutant.md)</code>.

```{r, eval=TRUE}
sim$add_mutant("A", list(E1 = list(duplication = 0.2, death = 0.05,
                                   E2 = 0.01),
                         E2 = list(duplication = 0.08, death = 0.01,
                                   E1 = 0.01, E3 = 0.01),
                         E3 = list(duplication = 0.08, death = 0.01,
                                   E2 = 0.01)))
```

The first parameter of the method is the name of the mutant to be added.
Instead, the second parameter of the method is a list of the rates to be set.

When the simulation has epigenetic states, the list must be a named list of
lists. Each name must be an epigenetic state, and the corresponding list
specifies the rates of the species. The names of this second type of list must
be among `"duplication"`, `"death"`, or any epigenetic state, while the values
represent the rates. When the name of a value is an epigenetic state, the value
itself is the rate of an epigenetic switch from the current state to the one
specified by the rate names. All missing rates are set to zero by default.

```{r}
# get the simulation mutants
sim$get_mutants()

# updated object (counts refer to number of cells of each species)
sim

# get the simulation rates that have been explicitly set
sim$get_rates()

# to get also the rates that haven't been explicitly set
sim$get_rates(complete = TRUE)
```

When, instead, the simulation has not epigenetic states, the list must be a
list of rates with names among `"duplication"` and `"death"`.
A mutant in a simulation without epigenetic states could be added as

```{r, eval=FALSE}
# not run
sim$add_mutant(name = "A", c(duplication = 0.2, death =  0.1))
```

To be able to simulate the model, an initial cell needs to be displaced
in the tissue.

```{r}
# add one cell of species A[E1] (mutant A in epi-state E1) in
# position (500, 500).
sim$place_cell("A[E1]", 500, 500)
```

### Visualisations

We can query the current state of the simulation and extract the
position of each cell in the tissue.

```{r}
# counts per species
sim$get_counts()

# cells position (one so far)
sim$get_cells()
```

This information can be plotted. Note that the tissue visualisation uses
hexagonal bins to avoid plotting delays when the simulation uses
thousands of cells.

```{r,  fig.height=4, fig.width=3, fig.alt=c("A pie-chart of the species cell count of a tumour consisting of a single cell.", "A representation of the tissue consisting of a single cell.")}
# pie-chart for counts
plot_state(sim)

# spatial distribution for the whole tissue (hexagonal bins)
plot_tissue(sim)
```

> **Note:** since the plots are done with `ggplot2` they can be assembled
> and customised.

## Species Evolution

There are 4 ways to let the simulation evolve:

-   advancing until the number of cells in a species reaches a given
    threshold $\theta > 0$;
-   advancing until a new time $t>0$ is reached;
-   advancing until a desired number of firings (of one particular
    event) has occurred;
-   advancing until a formula is not satisfied by the simulation
    status (_advanced topic_; see [this article](run_until.md)).

### Size-Based Simulation

We can run the simulation up to when we have $\theta > 0$; cells of
species `A[E1]`

```{r,  fig.height=4, fig.width=3, fig.alt=c("The tissue of a tumour with one mutant `A` and two species `A[E1]` and `A[E2]`. The species `A[E1]` consists of 500 cells.")}
# counts per species is now 0
sim$get_counts()

# run the simulation until the species "A[E1]" contains less than 500 cells
sim$run_up_to_size("A[E1]", 500)

# counts per species now reports 500 for A[E1]
sim$get_counts()

plot_tissue(sim)
```

### Firing-Based Simulation

The number of times each event has fired is accessible

```{r}
# get the number of fired event per species
sim$get_firings()
```

A small number of cell deaths have occurred in species `A[E2]` up to this
point, so we can simulate the system until there are 500 of them.

```{r}
sim$run_up_to_event("death", "A[E2]", 500)

# the row "death", for "A[E2]" now reports 1000
sim$get_firings()
```

```{r,  fig.height=4, fig.width=3, fig.alt=c("The tumour tissue after letting the simulation evolves until 100 deaths occurred to `A[E2]` cells.")}
# plot the tissue by using 200 bins
plot_tissue(sim, num_of_bins = 200)
```

### Clock-Based Simulation

It is also possible to take the current simulation clock as a reference,
and simulate further.

```{r}
# get the simulation clock
sim$get_clock()

# run the simulation for other 15 time units
sim$run_up_to_time(sim$get_clock() + 15)

# get again the simulation clock
sim$get_clock()
```

```{r,  fig.height=4, fig.width=3, fig.alt=c("The tumour tissue after letting the simulation evolves until 100 deaths occurred to `A[E2]` cells.")}
plot_tissue(sim)
```

##### Getting Cells (Advanced)

At this point, if we query the simulation, we will find more cells (we
use `dplyr` to ease result processing). For convenience, the getters
accept parameters to subset the tissue.

```{r, message=FALSE}
# load the package dplyr
library(dplyr)

# get the cells in the tissue at current simulation time
sim$get_cells() %>% head()

# get the cells in the tissue rectangular sample having
# [500,500] and [505,505] as lower and upper corners, respectively
sim$get_cells(c(500, 500), c(505, 505))  %>% head()

# get the cells in the tissue having epigenetic state "E2"
sim$get_cells(c("A", "B"), c("E2")) %>% head()

# get the cells in the tissue having epigenetic state "E2" and,
# at the same time, belonging to rectangular sample bounded by
# [500,500] and [505,505] as lower and upper corners, respectively
sim$get_cells(c(500, 500), c(505, 505), c("A", "B"), c("E2"))  %>% head()
```

### Evolving new species

ProCESS can select cells from the tissue randomly for each mutant, or
within a constrained tissue area.

```{r}
# random sampling from the whole tissue: it returns a cell having "A" as mutant
sim$choose_cell_in("A")

# calling it again may result in a different cell
sim$choose_cell_in("A")

# it can also sample among the cells of a particular species
sim$choose_cell_in("A[E2]")

# ... or multiple species/mutant
sim$choose_cell_in(c("A[E2]", "A[E3]"))

# constrain sampling in the tissue rectangular selection [500,550]x[350,450]
sim$choose_cell_in("A",  c(500, 350), c(550, 450))
```
It is also possible to constrain the selection to the cells being neighbors of
wild-type cells.

```{r}
# random sampling from the tumour border
sim$choose_border_cell_in("A")

# constrain sampling to rectangular selection [500,550]x[350,450]
sim$choose_border_cell_in(c("A[E2]", "A[E3]"),  c(500, 350), c(550, 450))
```

Choosing border cells is important when the border-driver growth model is
adopted (this is the default option). See [this article](growth_models.md)
for more details.

The ability to select a cell can be used to program the generation of new species,
mimicking new mutants that generate subclonal expansions.

Imagine we want to add a new mutant `B` with epi-states -- and therefore new
species `B[E1]`, `B[E2]`, `B[E3]`, and `B[E4]` -- as descending from `A`, we
need to:

-   locate one cell of mutant `A` in the tissue, which is where we
    will inject the new mutant;

-   add the specifics of mutant `B` (via <code>[TissueSimulation$add_mutant()](../reference/TissueSimulation-cash-add_mutant.md)</code>,
    as we did for `A`);

-   implement the change of the cell of mutant `A` to a cell of
    mutant `B`.

```{r}
# choose a border cell with mutant "A"
cell <- sim$choose_border_cell_in("A")
cell

# add mutant "B"
sim$add_mutant("B", list(E1 = list(duplication = 0.7, death = 0.05,
                                   E2 = 0.05),
                         E2 = list(duplication = 0.3, death = 0.1,
                                   E1 = 0.05),
                         E3 = list(death = 0.1, E2 = 0.05)))
```

Then we inject the cell, and let the simulation evolve until the number
of the cells in the species `B[E2]` is less than 10000.

```{r}
# mutate one of the children of `cell` into the species "B"
sim$mutate_progeny(cell, "B")

# generated event A -> B either in any of the added epi-states
sim$get_counts()

# evolve the simulation until "B[E2]" consists of less than 1e4 cells
sim$run_up_to_size("B[E2]", 10000)
```

At this point, we can inspect the tissue in more detail. It can help to
face the species to clearly appreciate the spatial diffusion of the
populations.

```{r,  fig.height=4, fig.width=3, fig.alt=c("The species count after the new mutant `B` arose.", "The tumour tissue after the new mutant `B` arose.")}
plot_state(sim)
plot_tissue(sim, num_of_bins = 250)
```

```{r,  fig.height=6, fig.width=6, fig.alt=c("The tumour tissue in which cells are split by species.")}
plot_tissue(sim, num_of_bins = 250) + ggplot2::facet_wrap(~species)
```

If, at this point in the simulation, we generate a new mutant `C` from
`A` in the rectangle $[450,500]\times [550, 650]$.

```{r}
# define evolutionary parameters
sim$add_mutant("C", list(E1 = list(duplication = 0.2, death = 0.1,
                                   E2 = 0.1),
                         E2 = list(duplication = 0.4, death = 0.01,
                                   E1 = 0.1)))

# choose a border cell in A[E1] or A[E2] in the rectangle [450,500]x[550,650]
cell <- sim$choose_border_cell_in(c("A[E1]", "A[E2]"),
                                  c(450, 550), c(500, 650))

sim$mutate_progeny(cell, "C")

sim$run_up_to_time(sim$get_clock() + 7)
```

```{r,  fig.height=4, fig.width=3, fig.alt=c("The species count after the third mutant `C` arose.", "The tumour tissue after the third mutant `C` mutant arose.")}
plot_state(sim)
plot_tissue(sim, num_of_bins = 250)
```

### Other Operations

##### Injection of Cells into a Tissue

In the tissue, we can manually inject multiple cells; all injected cells
can be retrieved.

```{r}
# now it will return just the initial cell
sim$get_added_cells()
```

##### Avoiding Drift

Any species with a non-zero death rate can become extinct
due to *drift*.

Drift makes it difficult to simulate and be confident about what species are
in the model. To facilitate the user,
[CLONES](https://github.com/albertocasagrande/CLONES) can avoid drift by
setting a *death activation level*. This value is the minimum number of cells
that enables cell death in a species: a cell of species $S$ can die if and
only if $S$ has reached the death activation level at least once during
the simulation.

This threshold applies to all species and is set to 1 by default.

```{r}
sim$death_activation_level

# change death activation level
sim$death_activation_level <- 50
```
