---
title: "Producing evolution videos"
output: rmarkdown::html_vignette
description: >
  How to produce a video showing tumour evolution
vignette: >
  %\VignetteIndexEntry{Producing evolution videos}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options:
  markdown:
    wrap: 72
---

```{r, include = FALSE}
req_pkgs <- c("av", "htmltools", "patchwork")
all_installed <- all(sapply(req_pkgs, requireNamespace, quietly = TRUE))

knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = all_installed
)
```

> *Disclaimer:* ProCESS/CLONES implements probability distributions using the
> C++11 random number distribution classes. Since the standard does not specify
> the underlying algorithms, their implementations are left to the compiler.
> Consequently, the simulation output depends on the compiler used to build
> [CLONES](https://github.com/albertocasagrande/CLONES), and the results
> reported in this article may differ from those obtained by the reader.

This article exclusively deals with producing videos of the tumour evolution.
Please, refer to [this article](tissue_simulation.md) for tumour evolution
modelling.

## Collecting data

In order to produce a video, we need to collect simulation snapshots during
tumour evolution. <code>[TissueSimulation](../reference/TissueSimulation_class.md)</code>
performs this task by default at the end of
<code>[place_cell()](../reference/TissueSimulation-cash-place_cell.md)</code>,
<code>[sample_cells()](../reference/TissueSimulation-cash-sample_cells.md)</code>,
<code>[mutate_progeny()](../reference/TissueSimulation-cash-mutate_progeny.md)</code>,
and <code>run_*()</code> calls.

Let us build a simulation and use
<code>[get_snapshot_info()](../reference/TissueSimulation-cash-get_snapshot_info.md)</code>
to inspect the collected snapshots.

```{r}
library("ProCESS")

# set the seed of the random number generator for repeatability
set.seed(0)

sim <- TissueSimulation(width = 600, height = 600)

# avoid drift
sim$death_activation_level <- 50

# add a mutant
sim$add_mutant("A", c(duplication = 0.12, death = 0.05))

# inspect the collected snapshots
sim$get_snapshot_info()

# place a cell and trigger the first snapshot
sim$place_cell("A", 300, 300)

# count the collected snapshots
nrow(sim$get_snapshot_info())

# let the model evolve and trigger the second snapshot
sim$run_up_to_size("A", 10)

# count the collected snapshots
nrow(sim$get_snapshot_info())

# add a further mutant
sim$add_mutant("B", c(duplication = 0.145, death = 0.06))

# one the "A" cell mutate into "B" and the third snapshot
# is collected
sim$mutate_progeny(sim$choose_border_cell_in("A"), "B")

# count the collected snapshots
nrow(sim$get_snapshot_info())

# let the model evolve and trigger the fourth snapshot
sim$run_up_to_size("B", 30)

# count the collected snapshots
nrow(sim$get_snapshot_info())

# collect a sample and get the fifth snapshot
sim$sample_cells("S1", c(145, 230), c(215, 300))

# count the collected snapshots
nrow(sim$get_snapshot_info())
```

The function <code>[build_snapshot_video()](../reference/build_snapshot_video.md)</code>,
which is exclusively available when the package
[av](https://CRAN.R-project.org/package=av) is installed, uses the
collected snapshot to build a video whose frames depicts the tissue as it was
when snapshots were collected.

The snippet

```{r, eval = FALSE}
build_snapshot_video(sim, quiet = TRUE)
```

produces the following video.

```{R, echo = FALSE}
output <- build_snapshot_video(sim, output_file = "video_1.mp4",
                               quiet = TRUE)

htmltools::tags$video(
  src = "video_1.mp4",
  type = "video/mp4",
  controls = TRUE,
  autoplay = TRUE,
  loop = TRUE,
  style = "max-width: 100%; height: auto;"
)
```

## Producing more frames

Whenever the snapshots collected by default are not enough to completely
represent the tumour evolution, user can set custom snapshot triggers.
ProCESS supports three types of triggers:
 - computation time triggers, e.g., collect a snapshot at least every 5 minutes
 - simulated time triggers, e.g., collect a snapshot at least every 5 simulated
   time units
 - number of tumour cells triggers, e.g., collect a snapshot at least every time
   the overall number of tumour cells increased by 1000 cells.

The <code>[TissueSimulation](../reference/TissueSimulation_class.md)</code> property
<code>[snapshot_triggers](../reference/TissueSimulation-cash-snapshot_triggers.md)</code>
reports the current custom triggers and can be used to set them.

```{R}
# see the current custom snapshot triggers
sim$snapshot_triggers

# take a new snapshot every 1000 new tumour cells
sim$snapshot_triggers <- list("number of cells" = 1000)

# get the current simulated time
sim$get_clock()

# count the snapshots collected before trigger changes
previous_snapshots <- nrow(sim$get_snapshot_info())

# let the model evolve until "B" consists of less than 2000 cells
sim$run_up_to_size("B", 2000)

# get the simulated time at each snapshot
snapshot_info <- sim$get_snapshot_info()

# show the number of tumour cells in each snapshot since the last default
# trigger
snapshot_info$cells[previous_snapshots:nrow(snapshot_info)]
```

<code>[snapshot_triggers](../reference/TissueSimulation-cash-snapshot_triggers.md)</code> can
also be used to reset custom triggers.

```{R}
# see the current custom snapshot triggers
sim$snapshot_triggers

# take a new snapshot every 1000 new tumour cells
sim$snapshot_triggers <- list("number of cells" = NULL, "clock interval" = 5)

# count the snapshots collected before trigger changes
previous_snapshots <- nrow(sim$get_snapshot_info())

# let the model evolve for further 20 simulated time units
sim$run_up_to_time(sim$get_clock() + 20)

# get the simulated time at each snapshot
snapshot_info <- sim$get_snapshot_info()

# show the simulated clock of each snapshot since trigger changes
snapshot_info$clock[previous_snapshots:nrow(snapshot_info)]
```
Custom snapshot triggers allow us to produce a more fluid evolution video.
For instance, let us consider the tumour model detailed in
[this article](four_mutants.md) and let us customise the snapshot triggers
just after creating the
<code>[TissueSimulation](../reference/TissueSimulation_class.md)</code> object as follows.

```{R, eval = FALSE}
library("ProCESS")

# set the seed of the random number generator for repeatability
set.seed(0)

sim <- TissueSimulation(width = 600, height = 600)

# collect a snapshot every 5 simulated time units
sim$snapshot_triggers <- list("clock interval" = 5)
```

The remaining part of the model specification is the same detailed
in [this article](four_mutants.md).

```{R, eval = FALSE}
# avoid drift
sim$death_activation_level <- 50

# add the mutant A
sim$add_mutant("A", c(duplication = 0.12, death = 0.05))

# place a cell in the tissue and simulate it until 10 cells
sim$place_cell("A", 300, 300)
sim$run_up_to_size("A", 10, quiet = TRUE)

# add the mutant B and let mutate a border cell of A in B
sim$add_mutant("B", c(duplication = 0.145, death = 0.06))
sim$mutate_progeny(sim$choose_border_cell_in("A"), "B")

# simulate the tissue up to 30 cells in B
sim$run_up_to_size("B", 30, quiet = TRUE)

# add the third mutant and let one cell of A mutate into C
sim$add_mutant("C", c(duplication = 0.15, death = 0.06))
sim$mutate_progeny(sim$choose_border_cell_in("A"), "C")

# simulate the tissue until C consists of 25000 cells
sim$run_up_to_size("C", 25000, quiet = TRUE)

# collect two samples
sim$sample_cells("S1", c(145, 230), c(215, 300))
sim$sample_cells("S2", c(350, 300), c(420, 370))

# add a further mutant and derive it from B
sim$add_mutant(name = "D", c(duplication = 0.8, death = 0.01))
sim$mutate_progeny(sim$choose_border_cell_in("B"), "D")

# let the tumour evolve until the mutant C and D cumulatively
# consist of 10000 cells
sim$run_until(sim$var("C") + sim$var("D") == 1e5, quiet = TRUE)

# collect the last sample
sim$sample_cells("S3", c(100, 400), c(170, 470))

# build the forest
sample_forest <- sim$get_sample_forests()
```

At the end of the evolution, by calling

```{R, eval = FALSE}
build_snapshot_video(sim, res = 300, width=1440, height=1024, framerate = 20)
```

we produce the following video.

```{R, echo = FALSE}
sim <- recover_simulation("four_mutants", quiet = TRUE)
sample_forest <- sim$get_sample_forest()

output <- build_snapshot_video(sim, res = 300, width=1440, height=1024,
                               framerate = 20, output_file = "video_2.mp4",
                               quiet = TRUE)

htmltools::tags$video(
  src = "video_2.mp4",
  type = "video/mp4",
  controls = TRUE,
  autoplay = TRUE,
  loop = TRUE,
  style = "max-width: 100%; height: auto;"
)
```

`build_snapshot_video` also supports pauses.

Users can specify pauses on events declaring the length of the pause
on specific sample or mutant rising or the generic event.

```{R, eval = FALSE}
pauses_on_event <- list(
    "mutant emerged" = list(

        # pause 4 seconds when "C" arises
        "C" = as.difftime(4, units="secs"),

        # pause 4 frames when "B" arises
        "B" = 4),

    # pause 3 seconds at each sample
    "sampling" = as.difftime(3, units="secs")
)

build_snapshot_video(sim, res = 300, width=1440, height=1024, framerate = 20
                     pauses_on_event = pauses_on_event)
```

```{R, echo = FALSE}
pauses_on_event <- list(
    "mutant emerged" = list(
        "C" = as.difftime(4, units="secs"),
        "B" = 4),

    "sampling" = as.difftime(3, units="secs")
)

output <- build_snapshot_video(sim, res = 300, width=1440, height=1024,
                               framerate = 20, output_file = "video_3.mp4",
                               pauses_on_event = pauses_on_event,
                               quiet = TRUE)

htmltools::tags$video(
  src = "video_3.mp4",
  type = "video/mp4",
  controls = TRUE,
  autoplay = TRUE,
  loop = TRUE,
  style = "max-width: 100%; height: auto;"
)
```

Pauses can also applied to specific frames.

```{R, eval = FALSE}
pauses_on_frame <- list(
    # pause 3 seconds on frame 101
    list(frame = 101, length = as.difftime(3, units="secs")),

    # pause 3 seconds on frame 102
    list(frame = 102, length = as.difftime(3, units="secs"))
)

build_snapshot_video(sim, res = 300, width=1440, height=1024, framerate = 20
                     pauses_on_frame = pauses_on_frame)
```

```{R, echo = FALSE}
pauses_on_frame <- list(
    list(frame = 101, length = as.difftime(3, units="secs")),
    list(frame = 102, length = as.difftime(3, units="secs"))
)

output <- build_snapshot_video(sim, res = 300, width=1440, height=1024,
                               framerate = 20, output_file = "video_4.mp4",
                               pauses_on_frame = pauses_on_frame,
                               quiet = TRUE)

htmltools::tags$video(
  src = "video_4.mp4",
  type = "video/mp4",
  controls = TRUE,
  autoplay = TRUE,
  loop = TRUE,
  style = "max-width: 100%; height: auto;"
)
```

## Customizing plots

<code>[build_snapshot_video()](../reference/build_snapshot_video.md)</code>
also supports
plot customisations. The parameter `plot_function` allows users to specify
their plotting function. The actual parameter should be a function that
takes in input a simulation and returns a plot. The plots of produced by
`plot_functions` taking in input the simulation snapshots are the frames of
the produced video.

For instance, we can produce a video focusing on the cells in a
rectangle. First of all, we need to define a focus function for the
cells in the rectangle (see, <code>[plot_tissue()](../reference/plot_tissue.md)</code>).

```{R}
# a focus function that highlights the cells in the rectangle
# [250,350]x[300,350]
in_rectangle <- function(cells) {
  (cells$position_x >= 250 & cells$position_x <= 350
   & cells$position_y >= 300 & cells$position_y <= 350)
}
```
Then, we may define the `plot_function`.

```{R}
# define a unique color map for all the frames
color_map <- get_species_colors(sim)

# this function will be called to plot each video frames
plot_in_rectangle <- function(snapshot) {
  plot_tissue(snapshot,
              color_map = color_map,
              plot_sample_region = FALSE,
              list_all_labels = TRUE,
              focus_function = in_rectangle)
}
```

The snippet

```{R, eval = FALSE}
build_snapshot_video(sim, res = 300, width=1440, height=1024,
                     framerate = 20, plot_function = plot_in_rectangle)
```

produces the following result.

```{R, echo = FALSE}
output <- build_snapshot_video(sim, res = 300, width=1440, height=1024,
                               framerate = 20, output_file = "video_5.mp4",
                               plot_function = plot_in_rectangle,
                               quiet = TRUE)

htmltools::tags$video(
  src = "video_5.mp4",
  type = "video/mp4",
  controls = TRUE,
  autoplay = TRUE,
  loop = TRUE,
  style = "max-width: 100%; height: auto;"
)
```


## Parallel frame generation

When the packages [furrr](https://CRAN.R-project.org/package=furrr) and
[progressr](https://CRAN.R-project.org/package=progressr) are installed in the 
system, the function
<code>[build_snapshot_video()](../reference/build_snapshot_video.md)</code>
uses them to parallelise frame
generations. However, ProCESS data structures are not serializable in R
and cannot be transparently passed to the
[furrr](https://CRAN.R-project.org/package=furrr)
processes. Because of that, all
ProCESS objects involved in the evaluation of `plot_function()`, with
the exception of the function parameter, must be saved to the disk
outside `plot_function()` and, then, reloaded inside it. In this way,
the objects will be local to each
[furrr](https://CRAN.R-project.org/package=furrr) process.

For instance, if we want to generate a video that focus on the cells
represented in the sample forest, we first need to save the sample
forest.

```{R}
sample_forest$save("forest.sff")
```

Then, we must define a `plot_function()` that loads its own copy
of the sample forest and use it.

```{R}
# this function will be called to plot each video frames
plot_in_forest <- function(snapshot) {
  # load the local copy of the forest
  local_forest <- load_forest("forest.sff", quiet = TRUE)

  # define a unique color map for all the frames
  color_map <- get_species_colors(local_forest)

  # a focus function that highlights the cells represented by
  # the sample forest
  in_sample_forest <- function(cells) {
    # sample_forest has been defined in four_mutants.md
    local_forest$represents_cell(cells$cell_id)
  }

  # plot the tissue using the shared color map 
  plot_tissue(snapshot,
              color_map = color_map,
              plot_sample_region = FALSE,
              list_all_labels = TRUE,
              focus_function = in_sample_forest)
}
```

Please, notice that `color_map` is a serialisable object and
is automatically passed to all frame generator processes.

Finally, the function call

```{R, eval = FALSE}
build_snapshot_video(sim, res = 300, width=1440, height=1024,
                     framerate = 20, plot_function = plot_in_forest)
```

produces the following video.

```{R, echo = FALSE}
output <- build_snapshot_video(sim, res = 300, width=1440, height=1024,
                               framerate = 20, output_file = "video_6.mp4",
                               plot_function = plot_in_forest,
                               quiet = TRUE)

htmltools::tags$video(
  src = "video_6.mp4",
  type = "video/mp4",
  controls = TRUE,
  autoplay = TRUE,
  loop = TRUE,
  style = "max-width: 100%; height: auto;"
)
```

The package [patchwork](https://CRAN.R-project.org/package=patchwork) can be
used to generate more complex videos.

```{R, eval = FALSE}
# this function will be called to plot each video frames
patchwork_function <- function(snapshot) {
  library(dplyr)

  # load the local copy of the forest
  local_forest <- load_forest("forest.sff", quiet = TRUE)

  # define a unique color map for all the frames
  color_map <- get_species_colors(local_forest)

  # a focus function that highlights the cells represented by
  # the sample forest
  in_sample_forest <- function(cells) {
    # sample_forest has been defined in four_mutants.md
    local_forest$represents_cell(cells$cell_id)
  }

  # plot the tissue using the shared color map 
  tissue_plot <- plot_tissue(snapshot,
                             color_map = color_map,
                             plot_sample_region = FALSE,
                             list_all_labels = TRUE,
                             focus_function = in_sample_forest)

  # get the snapshot simulated time
  clock <- snapshot$get_clock()

  # an alpha function that hides all nodes representing cells born after the
  # snapshot simulated time
  alpha_funct <- function(nodes) {
    nodes %>%
      dplyr::mutate(alpha = dplyr::case_when(birth_time <= clock ~ 1,
                                             TRUE ~ 0)) %>%
      dplyr::pull(alpha)
  }

  # generate the forest plot hiding the nodes representing cells that are
  # not born yet and removing the legends
  forest_plot <- plot_forest(local_forest, color_map = color_map,
                             alpha_function = alpha_funct) +
    ggplot2::guides(color = "none") +
    ggplot2::guides(shape = "none")

  events <- snapshot$get_just_occurred_events()

  if ("mutant emerged" %in% names(events)) {
    emerged_mutant <- events[["mutant emerged"]]
    cell <- snapshot$get_cells() %>%
      dplyr::filter(mutant == emerged_mutant)

    mutant_color <- color_map[[emerged_mutant]] 

    tissue_plot <- tissue_plot +
      ggrepel::geom_label_repel(data = cell,
                                ggplot2::aes(x = position_x, y = position_y,
                                             label = paste0("First cell in \"",
                                                            mutant,"\"")),
                                fill = mutant_color,
                                color = "black",
                                box.padding = 0.5, max.overlaps = Inf) +
      ggplot2::geom_point(data = cell,
                          ggplot2::aes(x = position_x, y = position_y),
                          color = mutant_color)

    forest_plot <- forest_plot +
      ggraph::geom_node_label(
        data = function(x) subset(x, name == as.character(cell$cell_id)),
        ggplot2::aes(label = paste0("First cell in \"",cell$mutant,"\"")),
        fill = mutant_color,
        segment.color = "black",
        segment.size = 0.5,
        min.segment.length = 0,
        #point.padding = 0,
        repel = TRUE
      ) +
      ggplot2::geom_point(
        data = function(x) subset(x, name == as.character(cell$cell_id)),
        ggplot2::aes(x = x, y = y),
        color = mutant_color,
        size = 0.8
      )

  } else if ("sampling" %in% names(events)) {
    sample_name <- events[["sampling"]]

    sample <- snapshot$get_samples_info() %>%
      dplyr::filter(name == sample_name)

    tissue_plot <- tissue_plot +
      ggplot2::annotate(
        "rect",
        xmin = sample$xmin, xmax = sample$xmax,
        ymin = sample$ymin, ymax = sample$ymax,
        fill = NA,
        color = "black"
      ) +
      ggplot2::annotate("text", x = (sample$xmin + sample$xmax) / 2,
                        y = (sample$ymin + sample$ymax) / 2,
                        label = sample_name, parse = TRUE)

    forest_plot <- forest_plot +
      ggplot2::geom_point(
        data = function(x) subset(x, sample == sample_name),
        ggplot2::aes(x = x, y = y),
        color = "black",
        size = 0.8
      )
  }
  
  # load the patchwork in each frame generator
  library(patchwork)
  
  # plot the tissue and the forest plots together
  (tissue_plot | forest_plot) +
    plot_layout(guides = "collect") &
    ggplot2::theme(legend.position = "bottom")
}

# define the pauses on events
pauses_on_event <- list(
  "mutant emerged" = as.difftime(3, units="secs"),
  "sampling" = as.difftime(2, units="secs")
)

build_snapshot_video(sim, width = 1820, height = 1024, framerate = 20,
                     res = 300, pauses_on_event = pauses_on_event,
                     plot_function = patchwork_function)
```

```{R, warning = FALSE, echo = FALSE}
# this function will be called to plot each video frames
patchwork_function <- function(snapshot) {
  library(dplyr)

  # load the local copy of the forest
  local_forest <- load_forest("forest.sff", quiet = TRUE)

  # define a unique color map for all the frames
  color_map <- get_species_colors(local_forest)

  # a focus function that highlights the cells represented by
  # the sample forest
  in_sample_forest <- function(cells) {
    # sample_forest has been defined in four_mutants.md
    local_forest$represents_cell(cells$cell_id)
  }

  # plot the tissue using the shared color map 
  tissue_plot <- plot_tissue(snapshot,
                             color_map = color_map,
                             plot_sample_region = FALSE,
                             list_all_labels = TRUE,
                             focus_function = in_sample_forest)

  # get the snapshot simulated time
  clock <- snapshot$get_clock()

  # an alpha function that hides all nodes representing cells born after the
  # snapshot simulated time
  alpha_funct <- function(nodes) {
    nodes %>%
      dplyr::mutate(alpha = dplyr::case_when(birth_time <= clock ~ 1,
                                             TRUE ~ 0)) %>%
      dplyr::pull(alpha)
  }

  # generate the forest plot hiding the nodes representing cells that are
  # not born yet and removing the legends
  forest_plot <- plot_forest(local_forest, color_map = color_map,
                             alpha_function = alpha_funct) +
    ggplot2::guides(color = "none") +
    ggplot2::guides(shape = "none")

  events <- snapshot$get_just_occurred_events()

  if ("mutant emerged" %in% names(events)) {
    emerged_mutant <- events[["mutant emerged"]]
    cell <- snapshot$get_cells() %>%
      dplyr::filter(mutant == emerged_mutant)

    mutant_color <- color_map[[emerged_mutant]] 

    tissue_plot <- tissue_plot +
      ggrepel::geom_label_repel(data = cell,
                                ggplot2::aes(x = position_x, y = position_y,
                                             label = paste0("First cell in \"",
                                                            mutant,"\"")),
                                fill = mutant_color,
                                color = "black",
                                box.padding = 0.5, max.overlaps = Inf) +
      ggplot2::geom_point(data = cell,
                          ggplot2::aes(x = position_x, y = position_y),
                          color = mutant_color)

    forest_plot <- forest_plot +
      ggraph::geom_node_label(
        data = function(x) subset(x, name == as.character(cell$cell_id)),
        ggplot2::aes(label = paste0("First cell in \"",cell$mutant,"\"")),
        fill = mutant_color,
        segment.color = "black",
        segment.size = 0.5,
        min.segment.length = 0,
        #point.padding = 0,
        repel = TRUE
      ) +
      ggplot2::geom_point(
        data = function(x) subset(x, name == as.character(cell$cell_id)),
        ggplot2::aes(x = x, y = y),
        color = mutant_color,
        size = 0.8
      )

  } else if ("sampling" %in% names(events)) {
    sample_name <- events[["sampling"]]

    sample <- snapshot$get_samples_info() %>%
      dplyr::filter(name == sample_name)

    tissue_plot <- tissue_plot +
      ggplot2::annotate(
        "rect",
        xmin = sample$xmin, xmax = sample$xmax,
        ymin = sample$ymin, ymax = sample$ymax,
        fill = NA,
        color = "black"
      ) +
      ggplot2::annotate("text", x = (sample$xmin + sample$xmax) / 2,
                        y = (sample$ymin + sample$ymax) / 2,
                        label = sample_name, parse = TRUE)
    
    forest_plot <- forest_plot +
      ggplot2::geom_point(
        data = function(x) subset(x, sample == sample_name),
        ggplot2::aes(x = x, y = y),
        color = "black",
        size = 0.8
      )
  }
  
  # load the patchwork in each frame generator
  library(patchwork)
  
  # plot the tissue and the forest plots together
  (tissue_plot | forest_plot) +
    plot_layout(guides = "collect") &
    ggplot2::theme(legend.position = "bottom")
}

# define the pauses on events
pauses_on_event <- list(
  "mutant emerged" = as.difftime(3, units="secs"),
  "sampling" = as.difftime(2, units="secs")
)

output_file <- build_snapshot_video(sim, output_file = "video_7.mp4",
                                    width = 1820, height = 1024,
                                    framerate = 20, res = 300,
                                    plot_function = patchwork_function,
                                    pauses_on_event = pauses_on_event,
                                    quiet = TRUE)

htmltools::tags$video(
  src = "video_7.mp4",
  type = "video/mp4",
  controls = TRUE,
  autoplay = TRUE,
  loop = TRUE,
  style = "max-width: 100%; height: auto;"
)
```

```{R, include = FALSE}
rm(sim)
rm(sample_forest)

unlink("four_mutants", recursive = TRUE)
unlink("forest.sff")
```
